## Lab 2: Functions and Environments


We are interested in the impact of different diets on chicken growth.
We have a dataset giving the weights of chickens on each of four different diets over the first several weeks of life in the `ChickWeight` dataset.

We can first look at the data and get a summary:

```{r}
data(ChickWeight)
head(ChickWeight)
summary(ChickWeight)
```

Questions:

- What type of object is `ChickWeight`?

- What types of data are the columns of `ChickWeight`? What happens when you try querying with either `typeof` or `class`? Do you get different results? Why?

We can also make a plot to see what's going on a little better.
Don't worry about `ggplot` for this class --- it's a nice plotting package, and if you take Exploratory Data Analysis you'll use it more, but you won't have to know about it for our class.
```{r}
library(ggplot2)
ggplot(ChickWeight, aes(x = Time, y = weight, color = Diet, group = Chick)) +
    geom_point() +
    geom_line() +
    facet_wrap(~ Diet)
```

Suppose we are interested in how much weight the chickens have put on after four days on diet 1.
We write the following code:
```{r}
time_zero_indices = ChickWeight$Time == 0 & ChickWeight$Diet == 1
end_time_indices = ChickWeight$Time == 4 & ChickWeight$Diet == 1
end_weight = mean(ChickWeight$weight[end_time_indices])
initial_weight = mean(ChickWeight$weight[time_zero_indices])
end_weight - initial_weight
```

We then decide that we might want to be able to compute this value for any diet and any end time, and so we would like to write a function to do so.
We try to do so in the following way:
```{r}
n_diets = length(unique(ChickWeight$Diet))
weight_gained_vector = rep(NA, n_diets)
set_weight_gained = function(diet, end_time) {
    time_zero_indices = ChickWeight$Time == 0 & ChickWeight$Diet == diet
    end_time_indices = ChickWeight$Time == end_time & ChickWeight$Diet == diet
    end_weight = mean(ChickWeight$weight[end_time_indices])
    initial_weight = mean(ChickWeight$weight[time_zero_indices])
    weight_gained_vector[diet] = end_weight - initial_weight
}
for(diet in 1:4) {
    set_weight_gained(diet = diet, end_time = 4)
}
```

Questions:

- What happens when you run the code above? Does the `weight_gained_vector` change after calling the `set_weight_gained` function? Why or why not?

- Map out the environments that are used when we call the `set_weight_gained` vector, and describe which environment each variable is in.

- Explain what the `diet = diet` part of the code in the for loop means. What is the formal argument and what is the actual argument?

The `ls()` function lists all of the variables present in the environment `ls` is called from.
We can use it to see what variables are present inside the function call by printing out the result from `ls` in the middle of the function.
Try that in the `set_weight_gained` function:
```{r}
set_weight_gained = function(diet, end_time) {
    time_zero_indices = ChickWeight$Time == 0 & ChickWeight$Diet == diet
    end_time_indices = ChickWeight$Time == end_time & ChickWeight$Diet == diet
    end_weight = mean(ChickWeight$weight[end_time_indices])
    initial_weight = mean(ChickWeight$weight[time_zero_indices])
    print(ls())
    weight_gained_vector[diet] = end_weight - initial_weight
}
set_weight_gained(diet = diet, end_time = 4)
```
We will see better tools for investigating the variables available in function environments when we talk about debugging, but for now this gives us a peek at what is happening when the function is called.

Your final task: Make a function that computes the average amount of weight gained by chickens on one of the diets after a given period of time.
That is, make a function that takes `diet` and `end_time`, like `set_weight_gained` did, but that returns the weight gained instead of trying to assign the output to a variable in the global environment.