```{r setup, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(fig.cap="", fig.width = 6, fig.height = 4, dpi=175,
               fig.path="lecture-2-fig/", warning = FALSE, cache=FALSE)
```





## Today

Reading: Matloff Chapter 7.1, 7.2, 2.9

Last time/finish up today: Data structures, so that we have something to work on.

This time: Flow control, so we can actually do things.




## Flow control and looping

- Conditionals

- Iteration

- Vectorization


## if statements

Syntax
```{r, eval = FALSE}
if (condition) {
    action1
} else {
    action2
}
```

. . .

So for example:

```{r}
weather <- "sunny"
if(weather == "rainy") {
    print("Take your umbrella!")
} else {
    print("No need for an umbrella today...")
}
```


-----

You can make more complicated conditions using either `else if` or
nested `if` statements:

. . .

```{r}
weather <- "cloudy"
if(weather == "rainy") {
    print("Take your umbrella!")
} else if (weather == "cloudy") {
    print("Think about taking your umbrella")
} else {
    print("No need for an umbrella today...")
}
```

-----

Some rules:

- `if` requires _one_ boolean, not a vector. It will throw a warning
if you give it a vector, but it will evaluated based on just the first
element of that vector.

- `else` is optional

- If the action is just one line, you don't need the braces (but you
should be consistent about this: choose a way you like and stick to
it).



## Combining booleans and lazy evaluation

We often want to combine conditions, which we can do with boolean
operations.

Like all other languages, R has AND and OR functions, but unlike some
other languages it has two of each.

- `&` and `&&` both mean AND.

- `|` and `||` both mean OR.

. . .

So for example:

```{r}
steak_type <- "med_rare"
temp <- 130
```
```{r}
if(steak_type == "rare" & temp > 115) {
    print("take your steak off!")
} else if(steak_type == "med_rare" & temp > 125) {
    print("take your steak off!")        
} else {
    print("you can keep cooking")
}
```

NB: As we'll see in two slides, `&` works here but it would be better
to use `&&`.

-----

Or, in not so dire a situation:

. . .

```{r}
steak_type <- "rare"
temp <- 110
```
```{r}
if(steak_type == "rare" && temp > 115) {
    print("take your steak off!")
} else if(steak_type == "med_rare" && temp > 125) {
    print("take your steak off!")        
} else {
    print("you can keep cooking")
}
```

-----

What is the difference between the two?

- Double operators give scalars (vectors of length 1), single
operators give vectors.

. . .

- If you call `&&` or `||` on a pair of vectors with length longer
than 1, the expression will be evaluated on the first element of the
vector.

. . .

- `&&` and `||` also support _lazy evaluation_.

-----

Lazy evaluation:

- `FALSE` followed by `&&` doesn't evaluate
the second expression.

- `TRUE` followed by `||` doesn't evaluate the second expression.

- This will occasionally make your code faster, so if you remember to
use `&&` and `||` for flow control and put the expressions that are
simpler to evaluate first, you will occasionally get performance improvements.


Try this on your computer. Which ones are fast and which ones are slow? Why?
```{r, eval = FALSE}
(FALSE && all(rep(1, 10^8) == 1))
(FALSE & all(rep(1, 10^8) == 1))
(all(rep(1, 10^8) == 1) && FALSE)
(all(rep(1, 10^8) == 1) & FALSE)
```

-----

Take-away:

- Use `&&` and `||` for flow control.

- Use `&` and `|` for operations on vectors.

- Put simpler operations first when using `&&` and `||`.

## Iteration

Two types

- For loops: You know how many iterations you need in advance.

- While loops: You'll know when to stop when you see it, but you don't
know in advance.

## For loops

Syntax:

```{r, eval = FALSE}
for(x in vector) {
    ...
}
```

Rules:

- `vector` is a vector

- `x` is a variable, which will be set first to `vector[1]`, then to
`vector[2]`, and so on, up to `vector[n]`, where `n` is the length of
`vector`.

- The actions inside `{` and `}` will be performed for each value of `x`.


-----

So for example:
```{r}
x <- 1:5
for(i in x) {
    print(i^2)
}
```

-----

As with all the other flow control elements, for loops can be nested.

. . .

We can use this to do something slightly more complicated:

```{r}
d <- 1:5
D <- matrix(NA, nrow = length(d), ncol = length(d))
D
```

```{r}
for(i in 1:nrow(D)) {
    for(j in 1:ncol(D)) {
        if(i == j) {
            D[i,j] <- d[i]
        } else {
            D[i,j] <- 0
        }
    }
}
D
```

-----

. . .

They can also be combined with the other flow control elements:

Don't worry about this part, just data setup:
```{r}
## install.packages("Lahman")
## install.packages("pacman")
library(Lahman)
library(pacman)
p_load(Lahman)
```

What the data looks like:
```{r}
head(People)
```

-----

. . .

And finally a for loop: What am I doing here?
```{r}
for(i in 1:nrow(People)) {
    if(!is.na(People$height[i]) && People$height[i] <= 62) {
        print(People[i,])
    }
}
```

Not a data problem:
[Eddie Gaedel](https://en.wikipedia.org/wiki/Eddie_Gaedel)

Question: does it matter whether we check for NA first? What could potentially happen if we check for height first instead?

## While loops

Syntax:

```{r, eval = FALSE}
while(condition) {
    ...
}
```

Rules:

- If `condition` is `TRUE`, the code inside `{` and `}` will be
evaluated.

- After the code inside `{` and `}` is evaluated, `condition` is
checked again, if it is still `TRUE`, we go again.

- This repeats until `condition` is FALSE.

-----

If you don't want your while loop to go forever, you have two options:

- The value of `condition` needs to eventually be set to `FALSE` by
the code inside `{` and `}`.

- You have a `break` statement inside the `{` `}` that eventually gets
you out of the loop.

-----

So for example, we could use a while loop to find the largest power of
2 less than 1000:

```{r}
x <- 2
while(x * 2 < 1000) {
    x <- x * 2
}
x
```

-----


Or for a slightly less silly example, we could use it to answer a
modified birthday problem.

Suppose we want to know how many classes filled with randomly selected
individuals we would have to attend before we found one where there
were at least two pairs of people with the same birthday.

We could go through the math, or we could get partway to an answer
with a while loop.

. . .

Here we draw sets of birthdays for classes of size 20, assuming that
there are 365 days in a year:

```{r}
days_in_year <- 365
class_size <- 20
num_classes <- 0
while(TRUE) {
    num_classes <- num_classes + 1
    birthdays <- sample(1:days_in_year, class_size, replace = TRUE)
    num_birthdays_per_day <- table(birthdays)
    days_with_match <- num_birthdays_per_day >= 2
    num_days_with_match <- sum(days_with_match)
    if(num_days_with_match >= 2) {
        break
    }
}
num_classes
```



. . .

Notes:

- The `while(TRUE)` with a `break` statement is a common idiom for
while loops.

- `break` can also be used in for loops.

## Vectorization

Most basic functions in R are _vectorized_, which means that they are
applied to vectors element-by-element.

- We already saw this with vector operations.

- Also true of many base functions.

```{r}
x <- rgamma(10, 1, .1)
x
log(x)
round(x)
floor(x)
```

-----

More on vectorization and its advantages later. Why vectorization?

- More readable code.

- Instead of writing _how_ you want the computer to perform the
computations, you tell the computer _what_ you want to do.

- Less typing = less of an opportunity to introduce bugs.

- Can be faster.


-----

Compare:

for-loop way of computing the floor of all the elements in the vector `x`:
```{r}
floor_of_x <- rep(NA, length(x)) ## pre-allocate a vector to hold our computations
for(i in 1:length(x)) {
    floor_of_x[i] <- floor(x[i])
}
floor_of_x
```

. . .

Versus the vectorized way:
```{r}
floor(x)
```

-----

## Vectorized conditionals


Suppose we want to plot the following function.

$$
f(x) = \begin{cases}
\frac{15}{16} (1 - x^2)^2 & |x| < 1\\
0 & \text{o.w.}
\end{cases}
$$


-----

. . .

Take 1:

```{r}
x <- seq(-2, 2, length = 200) ## a vector with the values at which we want to evaluate f
fx <- rep(NA, 200) ## pre-allocate a vector in which to store the values of f(x)
for(i in 1:200) {
    if(abs(x[i]) < 1) {
        fx[i] <- 15/16 * (1 - x[i]^2)^2
    } else {
        fx[i] <- 0
    }
}
plot(fx ~ x, type = 'l')
```

-----

## ifelse: Vectorized conditionals

`ifelse` syntax:
```{r, eval = FALSE}
ifelse(condition, yes, no)
```

Rules:

- `ifelse` returns a vector

- `condition` is a vector of Booleans.

- `yes` and `no` are vectors, should be of the same type.

. . .


`ifelse` goes element-by-element through `condition`, `yes`, and
`no`.

- The `i`th element of the output is `yes[i]` if `condition[i]` is
`TRUE`

- The `i`th element of the output is `no[i]` if `condition[i]` is `FALSE`


-----
. . .

Take 2:
```{r}
x <- seq(-2, 2, length.out = 200)
y <- ifelse(abs(x) < 1, 15/16 * (1 - x^2)^2, 0)
```

```{r}
plot(y ~ x, type = 'l')
```
