## Text representations and Data Frames

Reading: Matloff Chapter 11.1 (for strings), 6.1 (for factors), 5.1
(for data frames)


## The character type

. . .

- Strings have type `character` in R.

- No special type for single characters vs. strings:

```{r}
typeof('a')
typeof("ABC")
```

- As with the numeric type, single strings are vectors of length 1:

```{r}
length('a')
length("ABC")
```

- If you actually want the number of characters in a string, you need
the `nchar` function:

```{r}
nchar('a')
nchar("ABC")
```

## Creating strings

. . .

You can use either double or single quotes.

```{r}
"The Leopard"
'Burt Lancaster'
```

## Displaying strings

. . .

- `print` works.

- Sometimes slightly nicer to use `cat`, which prints the string to
the console instead of giving the representation with subscripts that
`print` does.

```{r}
print("The Leopard")
"The Leopard"
cat("The Leopard")
```

## Escape characters

The `\` is an _escape character_.

It tells R to interpret whatever comes after in a different way from normal.

There are two cases:

1. If the escape character comes before a character that would not usually be part of the string, the escape character indicates that the next character should be used literally, and not for its special meaning.

2. If the escape character comes before some kinds of real characters, the escape character modifies their behavior.

-----

. . .

As an example of case 1, the `"` or `'` in a string definition would usually mean that we come to the end of the string definition.

If you want a literal `"` or `'` in the string in that case, you need to escape it.
```{r}
cat("Giuseppe Tomasi di Lampedusa's \"The Leopard\"")
cat('Giuseppe Tomasi di Lampedusa\'s "The Leopard"')
```

-----

. . .

Another example of case 1: normally the `\` means to escape the next character. If you need a `\` in your string, you need to escape it:

```{r}
cat("We use the '\\' character to escape")
```

-----

## From last time: Greedy quantification

By default, quantifiers are greedy, meaning they match the longest
substring possible.

We can make them have the opposite behavior by modifying them with the
`?` character: in that case, they match the shortest substring
possible.

```{r}
regmatches("[i][j]", regexpr("\\[.*\\]", "[i][j]"))
regmatches("[i][j]", regexpr("\\[.*?\\]", "[i][j]"))
```



## Anchoring

- `^` (not inside square brackets) means that what comes after must be
at the start of a line.

- `$` means that what comes before must be at the end of a line.

- `\<` anchors to the beginning of a word.

- `\>` anchors to the end of a word.

- `\b` anchors to the boundary of words (beginning or ending)

- `\B` anchors to anywhere aside from the boundary

-----

. . .

How do we represent the anchors in R? 

Look at what happens if we try to create a string with the anchors:


```{r}
cat("\B", fill = TRUE)
```
```{r}
cat("\>", fill = TRUE)
```
```{r}
cat("\b", fill = TRUE)
```


So we need to escape the `\`:

```{r}
cat("\\B", fill = TRUE)
cat("\\>", fill = TRUE)
cat("\\b", fill = TRUE)
```

-----

. . .

```{r}
cat("\\<a", fill = TRUE)
grepl("\\<a", "hat at")
grepl("\\<a", "hat cat")
cat("\\bnana", fill = TRUE)
grepl("\\bnana", "bananas")
cat("\\Bnana", fill = TRUE)
grepl("\\Bnana", "bananas")
```


## String Manipulation

. . .

We can't use `[]` or `[[]]` to get at the internal parts of a string
because strings are atomic in R.

We need to use `substr`.

Syntax is `substr(string, start, stop)`:

```{r}
lancaster <- "Burt Lancaster"
substr(lancaster, 1, 4)
```

-----

. . .

`substr` vectorizes over the first argument (i.e. the `start` and
`stop` arguments will be expanded to match the length of the first argument):

```{r}
substr("Burt Lancaster", 1, 4)
substr("Burt Lancaster", c(1, 6), c(4, 14))
substr(rep("Burt Lancaster", 2), c(1, 6), c(4, 14))
substr(rep("Burt Lancaster", 2), 1, c(4, 14))
```

-----

. . .

We can use `substr` for replacement as well:

```{r}
lancaster
substr(lancaster, 1, 4) <- "Bill"
lancaster
```


## Combining strings

. . .

`paste` is the workhorse function for string combination.

Simplest way to use it:

`paste(s1, s2, ..., sn, sep)`

This will paste together `s1`, `s2`, up to `sn`, with `sep` in between
each one.

For example:

```{r}
paste("Chico", "Harpo", "Groucho", sep = ", ")
```

-----

. . .

The arguments to `paste` can be vectors, and the function is
vectorized so we get recycling:

```{r}
paste(c("Chico", "Harpo", "Groucho"), "Marx", sep = " ")
paste("Marx", c("Chico", "Harpo", "Groucho"), sep = ", ")
```

-----

. . .

Final argument: `collapse`.

Syntax `paste(vector, collapse)`

Will create _one string_ (vector of length 1, not the length of the
input vector) by pasting the elements of `vector` together, with the argument from `collapse`
in between them.

```{r}
paste(c("Chico", "Harpo", "Grouco"), collapse = ", ")
```

-----

. . .

Finaly, we can specify both `sep` and `collapse` together.

Think of this as first calling `paste` with `collapse = NULL`, then
calling `paste` with non-null collapse on the result:

```{r}
paste(c("Chico", "Harpo", "Groucho"), "Marx", sep = " ", collapse = " and ")
## note the equivalence:
marx.brothers = paste(c("Chico", "Harpo", "Groucho"), "Marx", sep = " ", collapse = NULL)
paste(marx.brothers, collapse = " and ")
```


## Splitting strings

. . .

Primariy function is `strsplit`.

Syntax: `strsplit(s, split)`

- `s` is a character vector (can have length greater than 1), and the
function vectorizes.

- `split` gives the string we want to split on: every element of `s`
will be split into pieces separated by `split`.

Given these parameters, what do you expect the output from `strsplit` to look
like? Which of the data structures we've seen so far can accommodate
everything we need?


-----

. . .

```{r}
split.brothers <- strsplit("Groucho and Harpo and Chico", "and")
typeof(split.brothers)
split.brothers

split.two.groups <- strsplit(c("Groucho and Harpo and Chico", "David and John"), "and")
typeof(split.two.groups)
split.two.groups
```

-----

. . .

A slightly more realistic example:

I have some fasta files that I'm going to use as input, perform some
manipulations on, and then write some output based on each one. I want
the output files to have the same prefix but have the extension `.txt`
instead of `.fasta`:

```{r}
file.names <- c("ighv_human.fasta", "ighd_mouse.fasta", "ighj_human.fasta", "ighv_mouse.fasta", "ighd_human.fasta", "ighj_mouse.fasta")
split.files <- strsplit(file.names, ".", fixed = TRUE)
output.names <- character(length = length(file.names))
for(i in 1:length(split.files)) {
    prefix <- split.files[[i]][[1]]
    output.names[i] <- paste(prefix, ".txt", sep = "")
}
output.names
```

## Factors

. . .

- A way of representing qualitative variables.

- Variable that takes string values, but we know in advance which
strings are valid.

-----

. . .

Factor creation:

- By default, when R reads in character data it will store it as a
factor.

- Manually with the factor function:

```{r}
factor(c("a", "b", "b", "z"))
```

-----

## Factors vs. strings

- Factors represented differently from strings internally.

- Factors are an integer vector plus a set of names corresponding to each of the levels.

-----

. . .

```{r}
state.name
typeof(state.name) ## typeof tells us about R's internal representation of the variable
state.name.fac <- as.factor(state.name)
typeof(state.name.fac) ## typeof tells us that from R's point of view, state.name.fac is a vector of integers 
class(state.name.fac) ## class tells us that the object is a factor
attributes(state.name.fac)
unclass(state.name.fac)
```

-----

. . .


This representation is a bit more parsimonious in general (but not
always...):

```{r}
object.size(state.name)
object.size(state.name.fac)
object.size(rep(state.name, 1000))
object.size(rep(state.name.fac, 1000))
```

-----

## Factor manipulation

Problems we might need to deal with:

- Too many levels

- Too few levels

- Levels with the wrong names

-----

. . .

Suppose California falls off into the ocean and Puerto Rico becomes a
state.

We need to change our state names to reflect the new state of affairs,
and we want to replace California with Puerto Rico.

```{r}
cal.index <- which(state.name.fac == "California")
state.name.fac[cal.index] = "Puerto Rico"
state.name.fac
```

Why does this happen?

-----

. . .

Let's try again:

```{r}
state.name.fac <- factor(state.name)
levels(state.name.fac) <- c(levels(state.name.fac), "Puerto Rico")
state.name.fac[cal.index] <- "Puerto Rico"
state.name.fac
```

-----

. . .

Then Puerto Rico decides to change its name to The People's Republic
of Puerto Rico, or The PR of PR, for its pleasing symmetry.

We need to modify our state names again. How can we do it?

```{r}
pr.index <- which(levels(state.name.fac) == "Puerto Rico")
levels(state.name.fac)[pr.index] <- "The PR of PR"
state.name.fac
```

-----

 . . .

We now have this extra level hanging around, and we would like to get
rid of it.

What can we do?

```{r}
cal.index <- which(levels(state.name.fac) == "California")
levels(state.name.fac) <- levels(state.name.fac)[-cal.index]
```

This is good behavior! It prevents us from messing up our factor
variables by mistake.

-----

. . .

How do we actually do this?

```{r}
droplevels(state.name.fac) ## droplevels function does what it says it does
factor(state.name.fac, levels = levels(state.name.fac)[-cal.index])
factor(state.name.fac)
```

## Ordered factors

No detail here, but for reference: ordered factors exist, are created with
the `ordered = TRUE` flag, some functions will have nicer default
behavior if you tell them that ordinal variables are ordinal.


## Data frames

Why do we need data frames?

- A common setup is that we have some samples, and some observations
on those samples. The observations are of different types: some are
numbers, and some are qualitative variables.

- We can't use a matrix because matrices
require all of the elements to be of the same type.
</br></br>
Note that this isn't just
an arbitrary rule about matrices: if we think about it from the
computer's point of view, the amount of space required for a matrix
containing all the same type of observations is fixed, but if the type
of the observation can change, we don't know how much space to
allocate.

- We need a data structure that mostly acts like a matrix
(rectangular, we can extract subsets easily, and so on), but that
allows for different data types in different columns.

-----

. . .

Data frames in R are lists with some extra rules:

- Each element of the list needs to be a vector

- Each of the vectors needs to be of the same length

This gives essentially a heterogeneous analog of a matrix.


## Data frame creation

```{r}
kids <- c("Jack", "Jill")
ages <- c(12, 10)
df <- data.frame(kids, ages)
df <- data.frame(Kids = kids, Ages = ages)
typeof(df)
class(df)
```



## Referring to elements of a data frame

Remember that data frames are lists, wich each element of the list
corresponding to one column.

This means we can use the `[[]]` notation to get a column:

```{r}
df[[1]]
```

. . .

We can also refer to portions of a data frame the same way we do for
matrices, with the `[]` notation:

```{r}
df[,1]
df[,1,drop = FALSE]
df[1,"Ages"]
```

. . .

Finally, we have a special operator, `$`, to get just one of the
columns of the data frame:

```{r}
df$Age
```

Pay attention to the return type: some subsetting operations give you
a data frame back, but some give you a vector.

## That's all, folks
