% Stat 470/670 Lecture 13: Level plots and contour plots for trivariate data
% Julia Fukuyama
% October 2, 2018

```{r setup, echo = FALSE}
library(knitr)
opts_chunk$set(fig.cap="", fig.width = 5, fig.height = 3, dpi=175, fig.path="lecture-13-fig/", warning = FALSE)
load("../../datasets/lattice.RData")
library(magrittr)
library(tidyverse)
library(ggplot2)
library(broom)
```

## Level plots and contour plots

Reading: Cleveland pp. 228-241, 245-248, 257-266, 270

Today: Two new kinds of plots

- Level plots

- Contour plots

We'll use them to plot

- Data

- Surfaces (e.g. output from a smoother)

Another theme that runs through the examples today: data is not collected at even intervals, and it helps to look in places where you have data.

## Level plots

We still have three continuous variables, two "explanatory" variables, $u$ and $v$, and one "response" variable, $y$.

Last week, we made coplots, with

- $y$ on the vertical axis,

- $u$ on the horizontal axis,

- Facets defined by cuts of $v$.

These are useful for finding interactions, and particularly useful when the interactions you find are simple.

## Level plot definition

A _level plot_ is very similar in form to a coplot, but the purpose is to visualize the response variable given both predictors at once.

In a level plot, we put

- $u$ on the vertical axis,

- $v$ on the horizontal axis,

- Define facets by cuts of $y$.

This allows us to see which regions of the explanatory variables correspond to similar values of the response variable.

## Galaxy data

Velocities of the NGC 7531 galaxy.

The data frame galaxy contains 323 measurements of velocity for locations in galaxy NGC 7531.

Variables are:

- `east.west`: Relative position on the east/west axis.

- `north.south`: Relative position on the north/south axis.

- `velocity`: Velocity of the galaxy in km/s.

We would like to model `velocity` as a function of position.


-----

Let's start out just plotting the explanatory variables to see where we have measurements.

```{r}
library(viridis)
load("../../datasets/lattice.RData")
summary(galaxy)
ggplot(galaxy, aes(x = east.west, y = north.south)) + geom_point(alpha=.5) + coord_fixed()
```

-----

Next, let's try a level plot of the data.

Remember that in a level plot, we plot the two explanatory variables on the horizontal and vertical axes, and we facet on intervals of the response variable.

```{r, fig.width=7.5}
## the cut_number command divides the range of a variable into
## intervals such that each interval has approximately the same
## number of observations
table(cut_number(galaxy$velocity, n = 10))
## we use cut_number to make the level plot/coplot: 
ggplot(galaxy, aes(x = east.west, y = north.south, color = velocity * 10^(-3))) + 
    geom_jitter(width = 0.5,  height = 0.5) +
    facet_wrap(~ cut_number(velocity * 10^(-3), n = 10), ncol = 5) +
    coord_fixed() +
    scale_color_viridis("Velocity, 10^3 km/s") + 
    labs(title = "Level plot of galaxy locations, faceted by velocity (expressed in 10^3 km/s)")
```

------

Compare the level plot with a plot in which velocity is indicated solely by color:

```{r}
ggplot(galaxy, aes(x = east.west, y = north.south, color = velocity * 10^(-3))) + 
    geom_jitter(width = 0.5,  height = 0.5) +
    coord_fixed() +
    scale_color_viridis("Velocity, 10^3 km/s") +
    labs(title = "Galaxy locations and velocity")
```


## Reviewing coplots and variable transformations

The galaxy dataset already has radial position and angle computed for us, but it's a useful exercise to do it yourself as well.

```{r}
galaxy %>%
    mutate(theta = atan(north.south / east.west) / (2 * pi) * 360) %>%
    mutate(rotation_to_horizontal = theta + 180 * (theta < 0)) %>%
    mutate(radial_position_recomputed = sqrt(north.south^2 + east.west^2)) %>%
    head
```

-----

Once we have the `angle` and `radial.position` variables, we can look at how velocity changes along each of the trajectories.

```{r}
ggplot(galaxy, aes(x = radial.position, y = velocity)) +
    geom_point() +
    stat_smooth(method = "loess") + 
    facet_wrap(~ angle, ncol = 4) +
    labs(title = "Galaxy position and velocity split by slit angle")
```

-------

Since there's nothing particularly special about the angles measured, and we expect that velocity is smooth not just along those trajectories but overall, we would prefer a 2d loess smooth.

```{r}
galaxy_loess = loess(velocity ~ east.west * north.south,
    data = galaxy, span = 0.25, 
    family = "symmetric", normalize = FALSE)
```

Why `normalize = FALSE`?

-----

Making the analogous coplot with loess fitted values, we see a similar but not identical smooth.

```{r, fig.width=7.5}
ggplot(augment(galaxy_loess, data = galaxy), aes(x = radial.position, y = velocity)) +
    geom_point() + 
    geom_line(aes(y = .fitted), color = "blue") +
    facet_wrap(~ angle, ncol = 4) +
    labs(title = "Galaxy position and velocity split by slit angle with fitted model")
```

## Level plots for fitted values

To make a level plot, we first need to get fitted values for the loess smoother at a grid of points:

```{r}
galaxy_grid = data.frame(expand.grid(
    east.west = seq(-25, 25, 0.5),
    north.south = seq(-45, 45, 0.5)))
galaxy_smoothed = augment(galaxy_loess, newdata = galaxy_grid)
```

------

Then we can make the level plot by cutting the response into bins:

```{r, fig.width=7.5}
ggplot(galaxy_smoothed, aes(x = east.west, y = north.south, fill = .fitted)) +
    geom_raster() + 
    coord_fixed() +
    facet_wrap(~ cut_number(.fitted * 10^-3, n = 20), ncol = 10) + 
    scale_fill_viridis()
```

------

Again, compare the level plot with one where the fitted value is indicated only by color:

```{r}
ggplot(galaxy_smoothed, aes(x = east.west, y = north.south, fill = .fitted)) +
    geom_raster() + 
    coord_fixed() +
    scale_fill_viridis()
```


## Contour plots

Definition: The $a$ _contour_ of a function $f(x,y)$ is the set $\{ (x, y) : f(x,y) = a \}$.

A _contour plot_ of $f(x,y)$ is a plot with $x$ and $y$ on the horizontal and vertical axes, and one or more contours of the function $f(x,y)$.

Note: Finding the contours is an inverse problem, and for most functions we want contours of we won't have anything available in closed form. In practice, contours are computed numerically.

------

An algorithm for finding the $a$ contour of a function $f$:

- Evaluate $f$ at a rectangular grid of points covering the region of study.

- Consider all the grid segments (line segments connecting two neighboring grid points) for which one endpoint is at greater than or equal to $a$ and the other endpoint is less than or equal to $a$. Note that if the function $f$ is continuous, the $a$ contour must cross the grid segment.

- Approximate the location where the $a$ contour crosses the grid segment by interpolation, and save the point.

- Draw contours by connecting the points (see pp. 242-244 for one way to do this).

------

Contour plot interpretation

- Closely spaced contours indicate large slope.

- Parallel, evenly-spaced contours indicate the surface is a plane.


## Contour plots for galaxy data

```{r}
ggplot(galaxy_smoothed, aes(x = east.west, y = north.south, fill = .fitted, z = .fitted)) +
    geom_raster() +
    geom_contour(bins = 20, color = "black") +
    coord_fixed() +
    scale_fill_viridis()
```


------

Again, compare with the plot with no contours and fitted value indicated only by color:

```{r}
ggplot(galaxy_smoothed, aes(x = east.west, y = north.south, fill = .fitted, z = .fitted)) +
    geom_raster() +
    coord_fixed() +
    scale_fill_viridis()
```


## Soil data

The `soil` dataset contains measurements of soil resistivity. The variables are:

- `northing`: The position along one axis (not actually north-south).

- `easting`: The position along the perpendicular axis (also not east-west).

- `resistivity`: The soil resistivity.


First let's look at where we have measures of resistivity:

```{r,fig.height=6}
ggplot(soil, aes(x = easting, y = northing)) + geom_point(size=.5) + coord_fixed()
```

-----

The plot above motivates the inclusion of the additional variables:

- `is.ns`: Was the measurement taken from a northing track?

- `track`: The ID of a track.

We can look at these two variables in a little more detail to see how tracks are defined:

```{r}
table(soil[,c("is.ns", "track")])
```

-----

As with the galaxy data, we can look at resistivity along tracks:

```{r, fig.width=7, fig.height=4}
ggplot(subset(soil, is.ns == TRUE), aes(x = northing, y = resistivity)) +
    geom_point(size=.5) + 
    facet_wrap(~track, ncol = 4)
```


```{r, fig.width=8, fig.height=4}
ggplot(subset(soil, is.ns == FALSE), aes(x = easting, y = resistivity)) +
    geom_point(size = 0.5) + 
    facet_wrap(~track, ncol = 8)
```

------

Because the variables are interchangeable, we would prefer a 2d loess fit:

```{r}
soil_lo = loess(resistivity ~ easting * northing, span = 0.25, data = soil, normalize = FALSE)
```

------

Once we have the fit, we evaluate it on a grid and make level plots and contour plots:
```{r, fig.height=6}
soil_grid = data.frame(expand.grid(easting = seq(0, 1.5, 0.01),
    northing = seq(0, 3.5, 0.01)))
soil_smoothed = augment(soil_lo, newdata = soil_grid)
ggplot(soil_smoothed,
       aes(x = easting, y = northing, z = .fitted, fill = .fitted)) +
    geom_raster() + 
    geom_contour(bins = 16, color = "black") +
    scale_fill_viridis() + 
    coord_fixed()
```

-----


```{r, fig.height=5.5}
ggplot(soil_smoothed,
       aes(x = easting, y = northing, fill = .fitted)) +
    geom_raster() +
    facet_wrap(~ cut_number(.fitted, n = 16), ncol = 8) + 
    scale_fill_viridis() + 
    coord_fixed()
```
