```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
theme_set(theme_bw())
set.seed(681)  # set once, here -- never inside a simulator function
```

Everything run in class today lives in this file (R) and its Python
counterpart. No separate lab this time.

# Section 1: Wrapping up the queueing lab

Same simulator as lecture 3, one customer at a time.

```{r queue-simulator}
simulate_queue <- function(n_customers, lambda, mu) {
  interarrival_duration <- numeric(n_customers)
  service_duration      <- numeric(n_customers)
  arrival_time          <- numeric(n_customers)
  service_start_time    <- numeric(n_customers)
  departure_time        <- numeric(n_customers)

  for (i in seq_len(n_customers)) {
    interarrival_duration[i] <- rexp(1, rate = lambda)
    if (i == 1) {
      arrival_time[i]       <- interarrival_duration[i]
      service_start_time[i] <- arrival_time[i]
    } else {
      arrival_time[i] <- arrival_time[i - 1] + interarrival_duration[i]
      service_start_time[i] <- max(arrival_time[i], departure_time[i - 1])
    }
    service_duration[i] <- rexp(1, rate = mu)
    departure_time[i]   <- service_start_time[i] + service_duration[i]
  }

  service_start_time - arrival_time  # waiting times
}
```

Hold `mu` fixed and let `lambda` creep up toward it -- the system gets
more utilized (higher traffic intensity $\rho = \lambda/\mu$):

```{r queue-blowup}
mu_fixed <- 1
lambda_seq <- seq(0.5, 0.95, by = 0.05)

mean_wait_at <- function(lambda, mu, n_customers = 1000, B = 100) {
  mean(replicate(B, mean(simulate_queue(n_customers, lambda, mu))))
}

mean_waits <- sapply(lambda_seq, mean_wait_at, mu = mu_fixed)

queue_df <- data.frame(lambda = lambda_seq, mean_wait = mean_waits)
ggplot(queue_df, aes(x = lambda, y = mean_wait)) +
  geom_line() +
  geom_point() +
  labs(x = expression(lambda), y = "mean waiting time",
       title = expression(paste("Mean wait vs. ", lambda, " (", mu, " = 1 fixed)"))) +
  theme_bw()
```

Nothing here estimates a parameter -- forward simulation alone already
told us something practically important: the system looks fine right up
until it doesn't, and the blowup is highly nonlinear. The same kind of
payoff shows up in epidemic thresholds, extinction probabilities in
population models, and staffing/capacity planning. Sometimes the
question really is just: *if the world worked this way, what would
happen?*


# Section 2: The SIR model, as another forward simulator

We didn't get to this last time. Three compartments:

$$
S \longrightarrow I \longrightarrow R
$$

Susceptible, Infected/infectious, Recovered/removed. Discrete time
steps, population size $N=S_t+I_t+R_t$ fixed, two binomial draws per
step:

$$
\Delta I_t \mid S_t, I_t \sim \operatorname{Binomial}\!\Big(S_t, 1-(1-\beta/N)^{I_t}\Big),
\qquad
\Delta R_t \mid I_t \sim \operatorname{Binomial}(I_t, \gamma),
$$
$$
S_{t+1} = S_t - \Delta I_t, \quad I_{t+1} = I_t + \Delta I_t - \Delta R_t, \quad R_{t+1} = R_t + \Delta R_t.
$$

$\beta$ is the transmission parameter, $\gamma$ the recovery parameter.
$R_0=\beta/\gamma$ is the basic reproduction number -- roughly, how many
new infections one infectious person generates while population immunity
is still negligible.

```{r sir-simulator}
simulate_sir <- function(n_steps, beta, gamma, S0, I0, R0) {
  N <- S0 + I0 + R0
  S <- integer(n_steps + 1); I <- integer(n_steps + 1); R <- integer(n_steps + 1)
  S[1] <- S0; I[1] <- I0; R[1] <- R0
  for (t in seq_len(n_steps)) {
    p_infect <- 1 - (1 - beta / N)^I[t]
    delta_I <- rbinom(1, S[t], p_infect)
    delta_R <- rbinom(1, I[t], gamma)
    S[t + 1] <- S[t] - delta_I
    I[t + 1] <- I[t] + delta_I - delta_R
    R[t + 1] <- R[t] + delta_R
  }
  data.frame(time = 0:n_steps, S = S, I = I, R = R)
}
```

## One simulated epidemic

```{r sir-single-run}
N <- 1000
I0 <- 20
gamma_known <- 0.15
n_steps <- 200

R0_illustrative <- 1.5
beta_illustrative <- R0_illustrative * gamma_known

sim1 <- simulate_sir(n_steps, beta_illustrative, gamma_known, N - I0, I0, 0)

sim1_long <- data.frame(
  time = rep(sim1$time, 3),
  count = c(sim1$S, sim1$I, sim1$R),
  compartment = factor(rep(c("S", "I", "R"), each = nrow(sim1)), levels = c("S", "I", "R"))
)

ggplot(sim1_long, aes(x = time, y = count, color = compartment)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(S = "steelblue", I = "firebrick", R = "darkgreen")) +
  labs(title = "One simulated epidemic (R0 = 1.5)", color = NULL) +
  theme_bw()
```

Is this *the* prediction of the model at $R_0=1.5$? Not exactly -- it's
one possible realization of what a stochastic model predicts:
$Y^{\text{sim}} \sim p(\cdot \mid \theta)$.

## Different parameters, different behavior

Two different $(\beta,\gamma)$ pairs:

```{r sir-two-parameter-trajectories}
sim_low  <- simulate_sir(n_steps, 0.225, gamma_known, N - I0, I0, 0)   # R0 = 1.5
sim_high <- simulate_sir(n_steps, 0.9,   gamma_known, N - I0, I0, 0)   # R0 = 6

compare_df <- data.frame(
  time = c(sim_low$time, sim_high$time),
  I = c(sim_low$I, sim_high$I),
  scenario = rep(c("beta=0.225, gamma=0.15 (R0=1.5)", "beta=0.9, gamma=0.15 (R0=6)"),
                  each = nrow(sim_low))
)

ggplot(compare_df, aes(x = time, y = I)) +
  geom_line(color = "firebrick") +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(y = "I(t)") +
  theme_bw()
```

Change the parameters and the whole trajectory changes shape -- size,
timing, speed. That's promising for fitting later. But eyeballing whole
trajectories doesn't scale to real data. We need a few numbers that
track these changes systematically.

## How do we measure "looks like"?

We want to find parameters that make simulated data resemble observed
data. How do we measure resemblance? For today, a first idea: hand-pick
summary statistics that seem to vary with the parameters.

Two candidates for today: **final epidemic size** (how many people were
ever infected) and **early growth rate** (slope of $\log I(t)$ over the
first few steps). Generically, $s(Y)$.

```{r sir-summary-fn}
window <- 8  # early-growth-rate fitting window

sir_summary_once <- function(R0) {
  beta <- R0 * gamma_known
  sim <- simulate_sir(n_steps, beta, gamma_known, N - I0, I0, 0)
  cum_infected <- N - sim$S
  final_size <- tail(cum_infected, 1)
  I_early <- sim$I[1:(window + 1)]
  t_early <- sim$time[1:(window + 1)]
  growth_rate <- unname(coef(lm(log(pmax(I_early, 1)) ~ t_early))[2])
  c(final_size = final_size, growth_rate = growth_rate)
}
```

If the simulated dataset is random, its summary is random too.

## From two parameters to one

For the systematic check that follows, fix $\gamma$ (say, known from
clinical data on infectious duration) and vary only $R_0=\beta/\gamma$.
One unknown parameter instead of two -- simpler to explore now, and
simpler to fit later.

Does a candidate summary actually help distinguish different values of
$R_0$? Look at its distribution across a range of $R_0$'s.

```{r sir-across-r0-setup}
R0_display <- c(2, 3, 4, 5, 6, 7, 8)
set.seed(13)
growth_by_r0 <- sapply(R0_display, function(r) replicate(150, sir_summary_once(r)["growth_rate"]))
set.seed(13)
final_by_r0  <- sapply(R0_display, function(r) replicate(150, sir_summary_once(r)["final_size"]))
```

## A useful summary: early growth rate

```{r sir-growth-boxplot}
growth_df <- data.frame(R0 = rep(R0_display, each = 150), growth_rate = as.vector(growth_by_r0))

ggplot(growth_df, aes(x = factor(R0), y = growth_rate)) +
  geom_boxplot() +
  labs(x = expression(R[0]), y = "early growth rate",
       title = "Growth rate: centers move, little overlap") +
  theme_bw()
```

## A weaker summary: final epidemic size

```{r sir-final-boxplot}
final_df <- data.frame(R0 = rep(R0_display, each = 150), final_size = as.vector(final_by_r0))

ggplot(final_df, aes(x = factor(R0), y = final_size)) +
  geom_boxplot() +
  labs(x = expression(R[0]), y = "final size",
       title = "Final size: centers move, then flatten and overlap") +
  theme_bw()
```

If you observed one epidemic, which statistic would tell you more about
$R_0$? Growth rate keeps separating across the whole range; final size
stops moving once most of the population is going to be infected either
way.

# Section 3: Turn the simulator around

We now have a good sense of how this simulator behaves -- which
summaries move, which don't. Time to use that to fit parameters to real
data.

Forward: $R_0 \to$ simulate epidemic $\to s(Y)$. Now reverse it:
$s(Y_{\text{obs}}) \to$ what $R_0$ could have produced this?

The algorithm you'd invent yourself:

1. choose a candidate $R_0$;
2. simulate many epidemics;
3. compute the same summary for each;
4. average the simulated summaries;
5. compare with the observed summary;
6. try another $R_0$.

Which candidate makes the simulated summaries look most like the
observed one?


# Section 4: Formalize SMM

$$
s_{\text{obs}} = s(Y_{\text{obs}}), \qquad
\hat m_B(\theta) = \frac1B\sum_{b=1}^B s(Y_\theta^{(b)}) \approx E_\theta[s(Y)].
$$

For one summary:
$$
Q(\theta) = \big[s_{\text{obs}} - \hat m_B(\theta)\big]^2,
\qquad
\hat\theta = \arg\min_\theta Q(\theta).
$$

## Objective-function plot: growth rate alone

```{r sir-objective-one-summary}
R0_true <- 6.0
R0_grid <- seq(4.0, 8.0, by = 0.1)
set.seed(4)
model_grid <- t(sapply(R0_grid, function(r) rowMeans(replicate(300, sir_summary_once(r)))))
colnames(model_grid) <- c("final_size", "growth_rate")

set.seed(5)
obs_illustrative <- sir_summary_once(R0_true)

Q_growth_only <- (model_grid[, "growth_rate"] - obs_illustrative["growth_rate"])^2
obj_df <- data.frame(R0 = R0_grid, Q = Q_growth_only)
R0_hat <- R0_grid[which.min(Q_growth_only)]

ggplot(obj_df, aes(x = R0, y = Q)) +
  geom_line() +
  geom_vline(aes(xintercept = R0_true, color = "true R0"), linetype = "dashed") +
  geom_vline(aes(xintercept = R0_hat, color = "estimated R0")) +
  scale_color_manual(name = NULL, values = c("true R0" = "blue", "estimated R0" = "red")) +
  labs(x = expression(R[0]), y = "Q(R0)", title = "Objective function, growth rate alone") +
  theme_bw()
```

**Simulated Method of Moments**: adjust parameters until simulated
summaries match observed summaries. "Moment" here just means a chosen
numerical summary of the data, not the classical statistical
method-of-moments machinery.


# Section 5: What if we use more than one summary?

$$
s(Y) = \begin{pmatrix}s_1(Y) \\ s_2(Y)\end{pmatrix},
\qquad
g(\theta) = s_{\text{obs}} - \hat m_B(\theta).
$$

How should two discrepancies become one number?

## First attempt: add squared discrepancies

$$
Q(\theta) = g_1(\theta)^2 + g_2(\theta)^2.
$$

This looks natural. It is not generally a good idea.

To do better we need an estimate of $\operatorname{Cov}(g(\theta))$. As a
shortcut today, we estimate it using simulations at the true $R_0$ (in
practice you don't know $R_0$ in advance: fit once with rough weighting,
estimate this near that estimate, then refit).

```{r sir-omega-estimate}
B_omega <- 500
set.seed(3)
at_true <- replicate(B_omega, sir_summary_once(R0_true))
Omega <- cov(t(at_true))
```

## Four ways to combine (or not combine) the two summaries

```{r sir-four-objectives-setup}
Omega_inv <- solve(Omega)

fit_once <- function(seed) {
  set.seed(seed)
  obs <- sir_summary_once(R0_true)  # one fresh "observed" epidemic
  g <- sweep(model_grid, 2, obs, "-")

  Q_strong     <- g[, "growth_rate"]^2
  Q_weak       <- g[, "final_size"]^2
  Q_unweighted <- rowSums(g^2)
  Q_weighted   <- apply(g, 1, function(row) as.numeric(t(row) %*% Omega_inv %*% row))

  c(strong     = R0_grid[which.min(Q_strong)],
    weak       = R0_grid[which.min(Q_weak)],
    unweighted = R0_grid[which.min(Q_unweighted)],
    weighted   = R0_grid[which.min(Q_weighted)])
}
```

## Discrepancies should be judged relative to ordinary variability

A discrepancy of 10 is huge if the statistic normally varies by 1, and
trivial if it normally varies by 100. For independent summaries:

$$
Q(\theta) = \frac{g_1(\theta)^2}{\sigma_1^2} + \frac{g_2(\theta)^2}{\sigma_2^2}.
$$

More generally, with possibly correlated summaries, use the full
covariance matrix:

$$
Q_W(\theta) = g(\theta)^\top W g(\theta), \qquad W \approx \operatorname{Cov}(g)^{-1}.
$$

```{r sir-omega-display}
Omega        # diagonal: variability of each summary; off-diagonal: correlation
```

Diagonal terms handle differences in variability; off-diagonal terms
account for correlation between summaries -- final size and growth rate
are not independent.

Weighting does **not** make every summary equally informative. It puts
discrepancies on an appropriate stochastic scale; how much a summary
then contributes still depends on how strongly its mean responds to the
parameter (the derivative) -- a property weighting can't manufacture.

## Does it actually help? (100 repeated fits, one fresh epidemic each)

```{r sir-repeated-study}
fits <- t(sapply(1:100, fit_once))

summarize <- function(x) c(bias = mean(x) - R0_true, rmse = sqrt(mean((x - R0_true)^2)))
round(t(apply(fits, 2, summarize)), 3)
```

Read this in order:

- **Weak alone** (final size) does clearly worse than **strong alone**
  (growth rate) -- matches the information calculation.
- **Unweighted** performs *identically* to weak alone: final size's raw
  scale (hundreds of people) so overwhelms growth rate's (fractions of a
  unit) that growth rate might as well not be in the objective at all.
- **Weighted** ($\hat\Omega^{-1}$) recovers the lost information -- it
  matches or beats even the best single summary.

A weak/noisy summary can hurt if combined carelessly; weighting keeps it
from having inappropriate influence.


# Section 6: The limits of weighting

But weighting is not enough on its own. What if a summary contains
almost no information about the parameter? What if two summaries contain
essentially the same information? What if our summaries miss an
important feature of the data entirely? No weighting matrix can rescue a
bad set of summaries.

$$
\boxed{
\theta \to \text{simulate many datasets} \to \text{calculate summaries} \to \text{estimate expected summaries}
}
$$
$$
\boxed{
\text{choose } \theta \text{ so simulated summaries resemble observed summaries}
}
$$

SMM turns a forward simulator into a parameter-estimation method.
