```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
theme_set(theme_bw())
set.seed(2026)  # set once, here -- never inside a simulator function
```

Code behind today's slides. There is no separate lab file for this
lecture.

# Section 1: Why not keep adding summaries?

## A worked example: adding a third summary

Recall estimating $(\beta,\gamma)$ in the SIR model. Using final size
and early growth rate gives an ok estimate of both. What if we add a
third summary, $I_2$, the raw case count on day 2?

```{r sir-third-summary-setup}
simulate_sir <- function(n_steps, beta, gamma, S0, I0, R0) {
  N <- S0 + I0 + R0
  S <- integer(n_steps + 1); I <- integer(n_steps + 1); R <- integer(n_steps + 1)
  S[1] <- S0; I[1] <- I0; R[1] <- R0
  for (t in seq_len(n_steps)) {
    p_infect <- 1 - (1 - beta / N)^I[t]
    delta_I <- rbinom(1, S[t], p_infect)
    delta_R <- rbinom(1, I[t], gamma)
    S[t + 1] <- S[t] - delta_I
    I[t + 1] <- I[t] + delta_I - delta_R
    R[t + 1] <- R[t] + delta_R
  }
  list(S = S, I = I, R = R)
}
N_sir <- 1000; I0_sir <- 20; n_steps_sir <- 60; window_sir <- 8; day2 <- 2
beta0 <- 0.30; gamma0 <- 0.10

summary_once <- function(beta, gamma) {
  sim <- simulate_sir(n_steps_sir, beta, gamma, N_sir - I0_sir, I0_sir, 0)
  final_size <- N_sir - tail(sim$S, 1)
  I_early <- sim$I[1:(window_sir + 1)]
  t_early <- 0:window_sir
  growth_rate <- unname(coef(lm(log(pmax(I_early, 1)) ~ t_early))[2])
  c(final_size = final_size, growth_rate = growth_rate, day2 = sim$I[day2 + 1])
}
```

## Is $I_2$ informative, redundant, or weak?

```{r sir-third-summary-diagnose, cache=TRUE}
reps_truth <- t(replicate(500, summary_once(beta0, gamma0)))
cor(reps_truth)          # I_2's correlation with the other two
apply(reps_truth, 2, sd)

# sensitivity: finite-difference approximation to d E[s]/d beta, evaluated
# at beta0 -- a 10% bump in beta, divided by the size of that bump (not
# just the raw shift in the summary's mean)
m0 <- colMeans(reps_truth)
delta_beta <- 0.1 * beta0
m1 <- colMeans(t(replicate(500, summary_once(beta0 + delta_beta, gamma0))))
sensitivity <- (m1 - m0) / delta_beta
sensitivity / apply(reps_truth, 2, sd)   # sensitivity-to-noise ratio
```

$I_2$ is not simply redundant with either of the other two summaries,
and its sensitivity-to-noise ratio is lower than growth rate's.

## Including $I_2$ degrades performance

Build a $(\beta,\gamma)$ grid, estimate $\hat\Omega^{-1}$ from a modest
simulation budget ($B=10$), and compare RMSE using 2 summaries vs. 3.

```{r sir-third-summary-grid, cache=TRUE}
beta_grid <- seq(0.15, 0.55, length.out = 17)
gamma_grid <- seq(0.05, 0.20, length.out = 17)
grid <- expand.grid(beta = beta_grid, gamma = gamma_grid)
B_grid <- 80
summary_avg <- function(beta, gamma, B) rowMeans(replicate(B, summary_once(beta, gamma)))
res <- t(mapply(function(b, g) summary_avg(b, g, B_grid), grid$beta, grid$gamma))
grid$final_size <- res[, "final_size"]; grid$growth_rate <- res[, "growth_rate"]; grid$day2 <- res[, "day2"]

estimate_bg <- function(obs, grid, Winv, mode) {
  if (mode == "two") {
    g <- cbind(obs["final_size"] - grid$final_size, obs["growth_rate"] - grid$growth_rate)
  } else {
    g <- cbind(obs["final_size"] - grid$final_size, obs["growth_rate"] - grid$growth_rate,
               obs["day2"] - grid$day2)
  }
  Q <- rowSums((g %*% Winv) * g)
  idx <- which.min(Q)
  c(beta = grid$beta[idx], gamma = grid$gamma[idx])
}

run_comparison <- function(B_omega, n_outer = 15, n_reps = 50) {
  rmse2 <- numeric(n_outer); rmse3 <- numeric(n_outer)
  for (o in seq_len(n_outer)) {
    reps <- t(replicate(B_omega, summary_once(beta0, gamma0)))
    Winv2 <- solve(cov(reps[, c("final_size", "growth_rate")]))
    Winv3 <- solve(cov(reps))
    err2 <- matrix(NA, n_reps, 2); err3 <- matrix(NA, n_reps, 2)
    for (i in seq_len(n_reps)) {
      obs <- summary_once(beta0, gamma0)
      e2 <- estimate_bg(obs, grid, Winv2, "two")
      e3 <- estimate_bg(obs, grid, Winv3, "three")
      err2[i, ] <- c(e2["beta"] - beta0, e2["gamma"] - gamma0)
      err3[i, ] <- c(e3["beta"] - beta0, e3["gamma"] - gamma0)
    }
    rmse2[o] <- sqrt(mean(rowSums(err2^2)))
    rmse3[o] <- sqrt(mean(rowSums(err3^2)))
  }
  list(rmse2 = rmse2, rmse3 = rmse3)
}

result_small_B <- run_comparison(B_omega = 10)
cat("mean RMSE, 2 summaries:", round(mean(result_small_B$rmse2), 4), "\n")
cat("mean RMSE, 3 summaries:", round(mean(result_small_B$rmse3), 4), "\n")
cat("fraction of trials where 3 summaries is worse:",
    mean(result_small_B$rmse3 > result_small_B$rmse2), "\n")
```

A genuinely valid, weak-but-not-redundant summary made the estimate
worse on average, and worse in most individual trials -- even though we
used the textbook-correct weighting rule, not $W=I$.

# Section 2: MMD as histogram distance

Running example for this section and the next: many independent
households, each starting with one index case. The recorded datum per
household is the total number infected, including the index case -- an
outcome in $\{1,\ldots,K\}$.

```{r household-sim}
household_sir <- function(n_steps, beta, gamma, S0, I0, R0) {
  N <- S0 + I0 + R0
  S <- integer(n_steps + 1); I <- integer(n_steps + 1); R <- integer(n_steps + 1)
  S[1] <- S0; I[1] <- I0; R[1] <- R0
  for (t in seq_len(n_steps)) {
    p_infect <- 1 - (1 - beta / N)^I[t]
    delta_I <- rbinom(1, S[t], p_infect)
    delta_R <- rbinom(1, I[t], gamma)
    S[t + 1] <- S[t] - delta_I
    I[t + 1] <- I[t] + delta_I - delta_R
    R[t + 1] <- R[t] + delta_R
  }
  list(S = S, I = I, R = R)
}
total_infected <- function(K, beta, gamma, n_steps = 60) {
  sim <- household_sir(n_steps, beta, gamma, K - 1, 1, 0)
  K - tail(sim$S, 1)
}

K <- 8
gamma0 <- 0.3
beta_obs  <- 2.5 * gamma0   # "observed": R0 = 2.5
beta_cand <- 1.5 * gamma0   # a candidate theta: R0 = 1.5
```

## Comparing observed and simulated data using histograms

```{r discrete-example}
n_obs <- 300
obs_counts <- replicate(n_obs, total_infected(K, beta_obs, gamma0))
sim_counts <- replicate(n_obs, total_infected(K, beta_cand, gamma0))
p_hat <- sapply(1:K, function(j) mean(obs_counts == j))
q_hat <- sapply(1:K, function(j) mean(sim_counts == j))
p_hat
q_hat
sum((p_hat - q_hat)^2)  # d^2 -- the simplest histogram distance
```

For comparison, two independent draws of 300 households at the same
$\theta$ as "observed" give a much smaller $d^2$ -- pure sampling noise:

```{r discrete-example-noise-floor}
sim_counts_same_theta <- replicate(n_obs, total_infected(K, beta_obs, gamma0))
q_hat_same <- sapply(1:K, function(j) mean(sim_counts_same_theta == j))
sum((p_hat - q_hat_same)^2)
```

```{r discrete-figure}
df_pq <- data.frame(
  category = factor(rep(1:K, 2)),
  prob = c(p_hat, q_hat),
  which = rep(c("observed  (p-hat)", "candidate theta  (q-hat)"), each = K)
)
ggplot(df_pq, aes(x = category, y = prob, fill = which)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(title = "Total infected per household (1-8 possible)", fill = NULL)
```

## Checking the general MMD definition against the histogram case

The general empirical definition of MMD, for any kernel $k$ and samples
$x_1,\ldots,x_n\sim P$, $y_1,\ldots,y_m\sim Q$:
$\widehat{\operatorname{MMD}}^2 = \frac{1}{n^2}\sum_{i,i'}k(x_i,x_{i'}) + \frac{1}{m^2}\sum_{j,j'}k(y_j,y_{j'}) - \frac{2}{nm}\sum_{i,j}k(x_i,y_j)$.
For the equality kernel, this reduces algebraically to $\sum_j(\hat p_j-\hat q_j)^2$:

```{r general-definition-check}
Kxx_avg <- mean(outer(obs_counts, obs_counts, `==`))
Kyy_avg <- mean(outer(sim_counts, sim_counts, `==`))
Kxy_avg <- mean(outer(obs_counts, sim_counts, `==`))
general_form <- Kxx_avg + Kyy_avg - 2 * Kxy_avg
direct_form  <- sum((p_hat - q_hat)^2)
c(general = general_form, direct = direct_form)
```

```{r equality-kernel-check}
outcomes <- 1:K
K_equality <- outer(outcomes, outcomes, function(a, b) as.numeric(a == b))
K_equality

as.numeric(t(p_hat - q_hat) %*% K_equality %*% (p_hat - q_hat))  # matches sum((p-q)^2) above
```

## SMM on the category frequencies

Define one summary per possible value of $X$ -- its sample proportion,
$s_j(X_1,\ldots,X_n) = \frac1n\sum_i \mathbf 1\{X_i=j\} = \hat p_j$ for
$j=1,\ldots,8$ -- and run ordinary SMM with this eight-dimensional
summary vector and $W=I$:

```{r smm-w-identity}
g_theta <- p_hat - q_hat
W_identity <- diag(length(g_theta))
Q_smm <- as.numeric(t(g_theta) %*% W_identity %*% g_theta)  # W = I
Q_smm
sum((p_hat - q_hat)^2)  # identical to the discrete MMD objective above
```

$Q(\theta) = g(\theta)^\top W g(\theta)$ with $W=I$ is exactly MMD
between observed and simulated data.

# Section 3: Smooth kernels

Same total-infected support, 1 through 8 -- these are ordered counts, so
a household with 2 people infected should plausibly count as more
similar to one with 3 than to one with 8. We use the Gaussian kernel as
the less-stringent alternative to the equality kernel.

```{r kernel-matrices}
K_eq <- outer(outcomes, outcomes, function(a, b) as.numeric(a == b))
h_k <- 1.5
K_smooth <- outer(outcomes, outcomes, function(a, b) exp(-(a - b)^2 / (2 * h_k^2)))

mat_to_df <- function(Km, label) {
  df <- expand.grid(x = outcomes, y = outcomes)
  df$k <- as.vector(Km)
  df$label <- label
  df
}
df_k <- rbind(mat_to_df(K_eq, "equality kernel"),
              mat_to_df(K_smooth, paste0("smooth kernel (h=", h_k, ")")))

ggplot(df_k, aes(x = x, y = y, fill = k)) +
  geom_raster() +
  facet_wrap(~label) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  coord_equal(expand = FALSE)
```

The same formula, a richer kernel: for a finite set of outcomes, any
kernel can be written as a similarity matrix $K$, and
$\operatorname{MMD}^2(P,Q) = (\mathbf p - \mathbf q)^\top K (\mathbf p - \mathbf q)$.
With $K=I$ (the equality kernel), this collapses back to
$\sum_j(p_j-q_j)^2$; with a smooth $K$, a mismatch that moves mass to a
neighboring category is weighted differently than one that moves it far
away.

```{r kernel-quadratic-form}
K_smooth_quad <- outer(outcomes, outcomes, function(a, b) exp(-(a - b)^2 / (2 * 1^2)))
as.numeric(t(p_hat - q_hat) %*% K_smooth_quad %*% (p_hat - q_hat))
```

# Section 5: A distribution mean and variance can't distinguish

```{r shape-family}
simulate_shape <- function(n, a) {
  S <- sample(c(-1, 1), n, replace = TRUE)
  eps <- rnorm(n, 0, sqrt(1 - a^2))
  a * S + eps
}

# confirm mean/variance invariance
for (a in c(0, 0.3, 0.6, 0.9)) {
  y <- simulate_shape(200000, a)
  cat("a =", a, " mean =", round(mean(y), 3), " var =", round(var(y), 3), "\n")
}
```

```{r shape-histograms}
a_show <- c(0, 0.3, 0.6, 0.9)
df_shape <- do.call(rbind, lapply(a_show, function(a) {
  data.frame(a = factor(paste0("a = ", a)), value = simulate_shape(4000, a))
}))
ggplot(df_shape, aes(x = value)) +
  geom_histogram(bins = 50, fill = "slateblue4", color = "white", boundary = 0) +
  facet_wrap(~a, nrow = 1) +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(title = "Same mean (0) and variance (1), increasingly separated shape")
```

## Two different methods for inferring $a$

`a0` is the (unknown, to the estimator) true value; both objectives are
built from one fixed observed sample and then evaluated by simulating at
each candidate `a`.

```{r mmd-function}
mmd2_gaussian <- function(x, y, h) {
  Kxx <- exp(-outer(x, x, "-")^2 / (2 * h^2))
  Kyy <- exp(-outer(y, y, "-")^2 / (2 * h^2))
  Kxy <- exp(-outer(x, y, "-")^2 / (2 * h^2))
  mean(Kxx) + mean(Kyy) - 2 * mean(Kxy)
}
```

```{r shape-objectives, cache=TRUE}
n_shape <- 500
a0 <- 0.7
x_obs <- simulate_shape(n_shape, a0)
a_grid <- seq(0, 0.95, by = 0.025)
B_rep <- 30
h_good <- 0.3

mmd_vals <- sapply(a_grid, function(a) {
  mean(sapply(1:B_rep, function(b) mmd2_gaussian(x_obs, simulate_shape(n_shape, a), h_good)))
})
obs_mean <- mean(x_obs); obs_var <- var(x_obs)
smm_vals <- sapply(a_grid, function(a) {
  mean(sapply(1:B_rep, function(b) {
    y <- simulate_shape(n_shape, a)
    (obs_mean - mean(y))^2 + (obs_var - var(y))^2
  }))
})

a_grid[which.min(mmd_vals)]  # should recover a0 = 0.7
a_grid[which.min(smm_vals)]  # should look essentially arbitrary
```

```{r shape-objectives-figure}
df_obj <- rbind(
  data.frame(a = a_grid, Q = smm_vals / max(smm_vals), method = "SMM: mean + variance"),
  data.frame(a = a_grid, Q = mmd_vals / max(mmd_vals), method = "MMD (Gaussian kernel)")
)
ggplot(df_obj, aes(x = a, y = Q, color = method)) +
  geom_line(linewidth = 1) + geom_point(size = 1.3) +
  geom_vline(xintercept = a0, linetype = "dashed", color = "gray40") +
  labs(title = "Only MMD localizes the true shape parameter",
       subtitle = paste0("dashed line: true a = ", a0),
       x = "candidate a", y = "objective (rescaled to own max)", color = NULL) +
  theme(legend.position = "bottom")
```

## So have we completely solved the problem?

The MMD curve above used one particular bandwidth. What happens at
other values of $h$?

```{r bandwidth-comparison, cache=TRUE}
h_values <- c(0.03, 0.3, 3)
mmd_by_h <- do.call(rbind, lapply(h_values, function(h) {
  vals <- sapply(a_grid, function(a) {
    mean(sapply(1:B_rep, function(b) mmd2_gaussian(x_obs, simulate_shape(n_shape, a), h)))
  })
  data.frame(a = a_grid, mmd = vals, h = factor(paste0("h = ", h), levels = paste0("h = ", h_values)))
}))

ggplot(mmd_by_h, aes(x = a, y = mmd)) +
  geom_line(color = "steelblue4", linewidth = 1) +
  geom_point(size = 1.1, color = "steelblue4") +
  geom_vline(xintercept = a0, linetype = "dashed", color = "gray40") +
  facet_wrap(~h, scales = "free_y", nrow = 1) +
  labs(title = "Bandwidth controls what MMD can see", x = "candidate a", y = "MMD^2")
```

Too small ($h=0.03$): only nearly-identical pairs register as similar at
all -- the true signal is there but noisier. Good ($h=0.3$): a more
convincing minimum. Too large ($h=3$): the kernel treats every pair of
points as almost equally similar, which washes out the signal.

# Section 7: Indirect inference with an SIR auxiliary model

Reuse the SIR simulator from Lectures 4-5. The auxiliary model is a
quadratic fit to $\log(I_t+1)$ over the first 20 steps -- level, growth,
curvature, in three numbers.

```{r sir-simulator}
N_pop <- 1000; I0 <- 20; gamma0 <- 0.10; window <- 20

aux_fit <- function(I_traj) {
  te <- 0:window
  y <- log(pmax(I_traj[1:(window + 1)], 1))
  unname(coef(lm(y ~ te + I(te^2))))
}

sim_example <- simulate_sir(window, 5 * gamma0, gamma0, N_pop - I0, I0, 0)
aux_fit(sim_example$I)
```

```{r aux-fit-figure}
te <- 0:window
y <- log(pmax(sim_example$I[1:(window + 1)], 1))
fit <- lm(y ~ te + I(te^2))
df_fit <- data.frame(t = te, log_I = y, fitted = fitted(fit))

ggplot(df_fit, aes(x = t)) +
  geom_point(aes(y = log_I)) +
  geom_line(aes(y = fitted), color = "firebrick", linewidth = 1) +
  labs(title = "Auxiliary model fit to one simulated epidemic",
       x = "t", y = "log(I_t + 1)")
```

$(\hat\alpha_0,\hat\alpha_1,\hat\alpha_2)$ would then be used as
summaries for this particular simulation -- fit the auxiliary model to
the one observed epidemic, average its fitted coefficients over many
simulated epidemics at each candidate $(\beta,\gamma)$, and move
$(\beta,\gamma)$ until that average resembles the observed fit.
