```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
theme_set(theme_bw())
```

Code behind the lecture 9 slides.

```{r B}
B <- 2000
```

# Section 1: A Normal location-scale model

We are measuring one quantity 500 times: some quality of a manufacturing
process, say the location of a pre-drilled hole in a piece of Ikea furniture.
Most of the time the hole is in the right place, but a small minority of the
time the machine messes up and you see a much wider spread.

```{r normal-data}
mu_true    <- 100
sigma_good <- 1
sigma_bad  <- 10
p_bad      <- 0.01
n_norm     <- 500

set.seed(11)
bad_reading <- rbinom(n_norm, 1, p_bad)
y_obs <- rnorm(n_norm, mu_true, ifelse(bad_reading == 1, sigma_bad, sigma_good))

sum(bad_reading)          # how many bad readings were drawn
```

## The observed measurements

```{r normal-observed-hist, fig.height=3.2}
ggplot(data.frame(y = y_obs), aes(x = y)) +
    geom_histogram(bins = 60, fill = "firebrick", colour = "white") +
    labs(x = "measurement", y = "count")
```

Almost all of the readings sit in a tight band. A handful are far away.

## The fitted model

Suppose we fit the model somehow and obtain $\hat\mu$ and $\hat\sigma$. It
does not matter for what follows how we got them.

```{r normal-fit}
mu_hat    <- mean(y_obs)
sigma_hat <- sqrt(mean((y_obs - mu_hat)^2))   # MLE, divides by n
c(mu_hat = mu_hat, sigma_hat = sigma_hat)
```

## Simulate and compare

```{r normal-hist, fig.height=3.2}
set.seed(12)
y_rep_one <- rnorm(n_norm, mu_hat, sigma_hat)

cmp <- rbind(data.frame(y = y_obs,     which = "observed"),
             data.frame(y = y_rep_one, which = "one simulation"))

ggplot(cmp, aes(x = y)) +
    geom_histogram(bins = 60, fill = "grey45", colour = "white") +
    facet_wrap(~ which, nrow = 1) +
    labs(x = "measurement", y = "count")
```

Two things are wrong at once: the simulated bulk is too wide, and the simulated
data never produce a reading anywhere near the far ones.

## Another way to see this: Q-Q plot

Sort both datasets. Plot the $k$-th smallest simulated value against the $k$-th
smallest observed value, for every $k$. The dashed line is where simulated
$=$ observed.

```{r normal-qq, fig.height=3.5}
qq_df <- data.frame(observed = sort(y_obs), simulated = sort(y_rep_one))

ggplot(qq_df, aes(x = observed, y = simulated)) +
    geom_abline(slope = 1, intercept = 0, colour = "steelblue",
                linetype = "dashed", linewidth = 1) +
    geom_point(alpha = 0.6, size = 1.6) +
    labs(x = "observed quantiles", y = "simulated quantiles")
```

## Is that just one unlucky simulation?

```{r normal-qq-many, fig.height=3.5}
set.seed(13)
qq_many <- do.call(rbind, lapply(1:50, function(b)
    data.frame(observed = sort(y_obs),
               simulated = sort(rnorm(n_norm, mu_hat, sigma_hat)),
               run = b)))

ggplot(qq_many, aes(x = observed, y = simulated)) +
    geom_abline(slope = 1, intercept = 0, colour = "steelblue",
                linetype = "dashed", linewidth = 1) +
    geom_line(aes(group = run), colour = "black", alpha = 0.15, linewidth = 0.4) +
    labs(x = "observed quantiles", y = "simulated quantiles")
```

## A one-number summary

How far from the middle is the most extreme reading?

$$
T(Y)=\max_i\left|Y_i-\operatorname{median}(Y)\right|.
$$

## The check in code

```{r normal-check}
outlier_distance <- function(y) max(abs(y - median(y)))

T_obs_tail <- outlier_distance(y_obs)

set.seed(2)
T_rep_tail <- replicate(B, outlier_distance(rnorm(n_norm, mu_hat, sigma_hat)))

c(observed   = T_obs_tail,
  rep_median = median(T_rep_tail),
  p_value    = (1 + sum(T_rep_tail >= T_obs_tail)) / (B + 1))
```

```{r normal-check-plot, fig.height=3.2}
ggplot(data.frame(T = T_rep_tail), aes(x = T)) +
    geom_histogram(bins = 40, fill = "grey75", colour = "white") +
    geom_vline(xintercept = T_obs_tail, colour = "firebrick", linewidth = 1.2) +
    labs(x = "max |y - median|", y = "replicate datasets")
```

The observed value is far outside anything the fitted model produces.

## What did this check tell us?

The failure is not evidence that $\mu$ or $\sigma$ was estimated imprecisely.
Changing $\mu$ moves the distribution and changing $\sigma$ stretches it.
Neither lets a Normal model produce a tight bulk *and* an occasional reading
thirteen units away from the middle.

# Section 2: Linear regression

## Reaction velocity and substrate concentration

The `Puromycin` data in R record reaction velocity at several substrate
concentrations. We use one treatment group and start with a linear model.

```{r puro-fit}
puro <- subset(Puromycin, state == "treated")
fit  <- lm(rate ~ conc, data = puro)
summary(fit)$coefficients
```

## The observed data

```{r puro-plot, fig.height=3.2}
ggplot(puro, aes(x = conc, y = rate)) +
    geom_point(colour = "firebrick", size = 2.8) +
    labs(x = "substrate concentration (ppm)", y = "reaction velocity")
```

Velocity rises with concentration, steeply at first and then much less so.

## What experiment should we imitate?

The concentrations were part of the experimental design, so we hold them fixed.
Each replicate draws new responses at exactly the observed `conc` values.

```{r puro-replicate}
mu_fitted <- fitted(fit)
sigma_lm  <- summary(fit)$sigma

simulate_response <- function() rnorm(nrow(puro), mu_fitted, sigma_lm)
```

## Simulate and compare

```{r puro-overlay, fig.height=3.4}
set.seed(31)
ovl <- rbind(
    data.frame(conc = puro$conc, rate = puro$rate,           which = "observed"),
    data.frame(conc = puro$conc, rate = simulate_response(), which = "one simulation")
)

ggplot(ovl, aes(x = conc, y = rate, colour = which, shape = which)) +
    geom_point(size = 2.8) +
    scale_colour_manual(values = c("observed" = "firebrick",
                                    "one simulation" = "grey35")) +
    labs(x = "substrate concentration (ppm)", y = "reaction velocity",
         colour = NULL, shape = NULL)
```

## Many simulations at once

```{r puro-violin, fig.height=3.6}
set.seed(32)
n_sim <- 500
sim_long <- data.frame(
    conc = rep(puro$conc, n_sim),
    rate = as.vector(replicate(n_sim, simulate_response()))
)

ggplot(sim_long, aes(x = conc, y = rate)) +
    geom_violin(aes(group = conc), width = 0.035,
                fill = "grey80", colour = "grey45") +
    geom_point(data = puro, colour = "firebrick", size = 2.8) +
    labs(x = "substrate concentration (ppm)", y = "reaction velocity")
```

Each grey shape is the distribution of simulated velocities at that
concentration.

## Averaging the runs at each concentration

```{r puro-violin-mean, fig.height=3.6}
obs_mean <- data.frame(
    conc = as.numeric(names(tapply(puro$rate, puro$conc, mean))),
    rate = as.numeric(tapply(puro$rate, puro$conc, mean))
)

set.seed(33)
sim_mean <- do.call(rbind, lapply(seq_len(n_sim), function(b) {
    y <- simulate_response()
    data.frame(conc = as.numeric(names(tapply(y, puro$conc, mean))),
               rate = as.numeric(tapply(y, puro$conc, mean)))
}))

ggplot(sim_mean, aes(x = conc, y = rate)) +
    geom_violin(aes(group = conc), width = 0.035,
                fill = "grey80", colour = "grey45") +
    geom_point(data = obs_mean, colour = "firebrick", size = 2.8) +
    labs(x = "substrate concentration (ppm)", y = "mean reaction velocity")
```

Each concentration was run twice and averaged, and the red dots are the average
of the observed data at each concentration.

## One potential discrepancy

Every concentration was run exactly twice, and so we can measure the amount of
noise using the absolute difference between the two observations at a given
concentration.

$$
T(Y)=\sum_{\text{concentrations }c}\left|Y_{c,1}-Y_{c,2}\right|.
$$

## The pairs check in code

```{r puro-pairs}
pair_spread <- function(y) sum(abs(tapply(y, puro$conc, diff)))

T_obs_pairs <- pair_spread(puro$rate)

set.seed(4)
T_rep_pairs <- replicate(B, pair_spread(simulate_response()))

# Lack of fit can only inflate sigma_hat, so this is a lower-tail check.
c(observed   = T_obs_pairs,
  rep_median = median(T_rep_pairs),
  p_value    = (1 + sum(T_rep_pairs <= T_obs_pairs)) / (B + 1))
```

```{r puro-pairs-plot, fig.height=3.5}
ggplot(data.frame(T = T_rep_pairs), aes(x = T)) +
    geom_histogram(bins = 40, fill = "grey75", colour = "white") +
    geom_vline(xintercept = T_obs_pairs, colour = "firebrick", linewidth = 1.2) +
    labs(x = "total spread within the replicate pairs", y = "replicate datasets")
```

## Two different ways to see one failure

The violin plots and this statistic detect the same defect in different ways:

* the violins show the true mean function bending where the line is straight;
* the pairs show the noise estimate inflated to cover that bend.

```{r puro-pure-error}
c(sigma_hat_from_line = sigma_lm,
  pure_error_sd = sqrt(sum(tapply(puro$rate, puro$conc, diff)^2) /
                       (2 * length(unique(puro$conc)))))
```

# Section 3: A single-server queue

## We have data and a fitted model

We observe one queue: the arrival times and the waiting times of a long run of
customers. Suppose we have fitted $\lambda$ and $\mu$ somehow. As before, how
we fitted them does not matter here.

```{r queue-sim}
queue_from_durations <- function(interarrival, service) {
    n <- length(interarrival)
    arrival   <- cumsum(interarrival)
    start     <- numeric(n)
    departure <- numeric(n)
    for (i in seq_len(n)) {
        start[i]     <- if (i == 1) arrival[i] else max(arrival[i], departure[i - 1])
        departure[i] <- start[i] + service[i]
    }
    data.frame(customer = seq_len(n), interarrival, service,
               arrival, start, departure, waiting = start - arrival)
}

simulate_queue <- function(n_customers, lambda, mu)
    queue_from_durations(rexp(n_customers, lambda), rexp(n_customers, mu))
```

```{r queue-data}
n_cust <- 1000
lambda_true <- 0.50

set.seed(5)
slow <- rbinom(n_cust, 1, 0.1)
service_obs <- ifelse(slow == 1, rexp(n_cust, rate = 1 / 4.6),
                                 rexp(n_cust, rate = 1 / 0.6))
interarrival_obs  <- rexp(n_cust, lambda_true)
queue_obs   <- queue_from_durations(interarrival_obs, service_obs)

lambda_hat <- 1 / mean(interarrival_obs)
mu_hat_q   <- 1 / mean(service_obs)
c(lambda_hat = lambda_hat, mu_hat = mu_hat_q, rho = lambda_hat / mu_hat_q)
```

## The observed waiting times

```{r queue-observed-hist, fig.height=3.2}
ggplot(data.frame(w = queue_obs$waiting), aes(x = w)) +
    geom_histogram(bins = 50, fill = "firebrick", colour = "white") +
    labs(x = "waiting time", y = "customers")
```

## Simulate and compare

Simulate one complete queue from the fitted model (same number of customers,
same fitted rates) and put its waiting times beside the observed ones.

```{r queue-hist-compare, fig.height=3.2}
set.seed(8)
queue_rep_one <- simulate_queue(n_cust, lambda_hat, mu_hat_q)

qcmp <- rbind(
    data.frame(w = queue_obs$waiting,     which = "observed"),
    data.frame(w = queue_rep_one$waiting, which = "one simulation")
)

ggplot(qcmp, aes(x = w, fill = which)) +
    geom_histogram(bins = 50, colour = "white") +
    facet_wrap(~ which, nrow = 1) +
    scale_fill_manual(values = c("observed" = "firebrick",
                                  "one simulation" = "grey45")) +
    guides(fill = "none") +
    labs(x = "waiting time", y = "customers")
```

## The queue check in code

```{r queue-wait}
mean_wait <- function(q) mean(q$waiting)

T_obs_wait <- mean_wait(queue_obs)
set.seed(7)
T_rep_wait <- replicate(B, mean_wait(simulate_queue(n_cust, lambda_hat, mu_hat_q)))

c(observed   = T_obs_wait,
  rep_median = median(T_rep_wait),
  p_value    = (1 + sum(T_rep_wait >= T_obs_wait)) / (B + 1))
```

```{r queue-wait-plot, fig.height=3.5}
ggplot(data.frame(T = T_rep_wait), aes(x = T)) +
    geom_histogram(bins = 40, fill = "grey75", colour = "white") +
    geom_vline(xintercept = T_obs_wait, colour = "firebrick", linewidth = 1.2) +
    labs(x = "mean waiting time", y = "replicate datasets")
```

## What did this check tell us?

* Mainly that the model is insufficient to explain waiting times.
* You would need to know a bit more about the model and play around with other
  possibilities to see how to fix it.
* In this case, it could be fixed by modifying the service-time distribution
  (the data were actually generated under a model where arrivals were from an
  exponential distribution but service times were from a distribution with
  heavier tails).

# What we saw

* Uncertainty quantification and model checking answer different questions.
* A generative model makes predictions beyond the quantities used to fit it.
* We check those predictions by simulating complete datasets and calculating
  the same features in observed and replicated data.
* A failed check is most valuable when it points toward a specific weakness in
  the model and a plausible revision.
