# Three possible coins

Simulate $\theta\in\{1/4,1/2,3/4\}$ and two coin flips.

```{r three-coin-simulator}
set.seed(681)
simulate_three_coin <- function(prior_probs = c(1/3, 1/3, 1/3)) {
  theta <- sample(c(0.25, 0.5, 0.75),
                  size = 1, prob = prior_probs)
  X <- rbinom(2, size = 1,
              prob = theta)
  return(list(theta = theta, X = X))
}
```

Keep only HH, coded as $X=(1,1)$.

```{r three-coin-rejection}
rejection_sample_three_coin <-
  function(prior_probs = c(1/3, 1/3, 1/3)) {
    while (TRUE) {
      run <- simulate_three_coin(prior_probs)
      if (all(run$X == c(1, 1)))
        return(run$theta)
    }
  }
rejection_sample_three_coin()
```

Estimate the posterior probabilities from accepted draws.

```{r three-coin-posterior}
draws <- replicate(5000,
  rejection_sample_three_coin())

# Fraction of accepted draws
# at each value:
table(draws) / 5000
```

Repeat with prior probabilities $(0.6,0.3,0.1)$.

```{r three-coin-new-prior}
new_prior <- c(0.6, 0.3, 0.1)
new_draws <- replicate(5000,
  rejection_sample_three_coin(new_prior))
table(factor(new_draws,
  levels = c(0.25, 0.5, 0.75))) / 5000
```

# Continuous prior for a coin

Simulate $\theta\sim\operatorname{Uniform}(0,1)$ and six flips.

```{r continuous-coin-simulator}
set.seed(681)
simulate_continuous_coin <- function() {
  theta <- runif(1)
  X <- rbinom(6, size = 1,
              prob = theta)
  return(list(theta = theta, X = X))
}
```

Keep the exact observed sequence HHTHHT.

```{r continuous-coin-rejection}
rejection_sample_continuous_coin <-
  function() {
    while (TRUE) {
      run <- simulate_continuous_coin()
      if (all(run$X == c(1, 1, 0, 1, 1, 0)))
        return(run$theta)
    }
  }
rejection_sample_continuous_coin()
```

Estimate the posterior after matching the complete sequence.

```{r continuous-coin-posterior}
draws <- replicate(2000,
  rejection_sample_continuous_coin())
mean(draws)
hist(draws, breaks = 40,
     xlim = c(0, 1),
     xlab = "theta", main = "")
```

Keep any sequence with four heads instead.

```{r continuous-coin-count}
rejection_sample_coin_count <-
  function() {
    while (TRUE) {
      run <- simulate_continuous_coin()
      if (sum(run$X) == 4)
        return(run$theta)
    }
  }
draws_count <- replicate(2000,
  rejection_sample_coin_count())
```

# Dependent rainy days

Simulate six days; each day after the first persists with probability $1/2$ or redraws using $\theta$.

```{r rain-simulator}
set.seed(681)
simulate_rain <- function() {
  theta <- runif(1)
  X <- integer(6)
  X[1] <- rbinom(1, size = 1, prob = theta)
  for (day in 2:6) {
    if (runif(1) < 0.5) {
      X[day] <- X[day - 1]                         # weather persists
    } else {
      X[day] <- rbinom(1, size = 1, prob = theta)   # new state
    }
  }
  return(list(theta = theta, X = X))
}
```

Use one rejection sampler with a condition $E$ on the six-day sequence.

```{r rain-rejection}
rejection_sample_rain <-
  function(E) {
    while (TRUE) {
      run <- simulate_rain()
      if (E(run$X))
        return(run$theta)
    }
  }
```

Match the complete sequences DRRRRD and RDRDRR.

```{r rain-exact-patterns}
E_DRRRRD <- function(X)
  all(X == c(0, 1, 1, 1, 1, 0))
E_RDRDRR <- function(X)
  all(X == c(1, 0, 1, 0, 1, 1))

rejection_sample_rain(E_RDRDRR)
```

Compare posterior means for the two complete sequences.

```{r rain-exact-posteriors}
draws <- function(E)
  replicate(2000,
    rejection_sample_rain(E))

mean(draws(E_DRRRRD))
mean(draws(E_RDRDRR))
```

Condition on four rainy days, regardless of order.

```{r rain-count-posterior}
E_count <- function(X)
  sum(X) == 4
count_draws <- draws(E_count)
mean(count_draws)
```

