# First example: normal location model

$\theta \sim \text{Uniform}[-3,3]$, $X \mid \theta \sim N(\theta, 1)$. Suppose we observe $x_{\text{obs}} = 0.8$.

```{r normal-simulator}
set.seed(681)
simulate_normal <- function() {
  theta <- runif(1, -3, 3)
  X <- rnorm(1, mean = theta,
             sd = 1)
  return(list(theta = theta,
              X = X))
}
```

Try rejection sampling.

```{r normal-rejection}
rejection_sample_normal <-
  function(x_obs = 0.8) {
    while (TRUE) {
      run <- simulate_normal()
      if (run$X == x_obs)
        return(run$theta)
    }
  }

X <- replicate(100000,
  simulate_normal()$X)
sum(X == 0.8)   # 0
```

Doesn't work! $P(X = 0.8) = 0$, so `rejection_sample_normal()` never returns. (Don't run it.)

# Approximate Bayesian computation

Keep $\theta^*$ if $d(S(X^*), S(x_{\text{obs}})) < \epsilon$. Take $S(x) = x$ and $d(x,y) = |x - y|$.

```{r abc-normal}
abc_sample_normal <-
  function(eps, x_obs = 0.8) {
    while (TRUE) {
      run <- simulate_normal()
      if (abs(run$X - x_obs) < eps)
        return(run$theta)
    }
  }
```

# What does $\epsilon$ do?

Compare $\epsilon = 1.5$, $0.2$, $0.05$. The orange line is the exact posterior.

```{r tolerance-comparison, fig.width = 10, fig.height = 3.5}
eps_values <- c(1.5, 0.2, 0.05)
draws <- lapply(eps_values, function(eps)
  replicate(2000, abc_sample_normal(eps)))

grid <- seq(-3, 3, length.out = 1001)
exact <- dnorm(grid, 0.8, 1) / (pnorm(3, 0.8, 1) - pnorm(-3, 0.8, 1))

par(mfrow = c(1, 3))
for (j in 1:3) {
  hist(draws[[j]], breaks = seq(-3, 3, by = 0.2), freq = FALSE,
       ylim = c(0, 0.5), xlab = "theta",
       main = paste("Tolerance", eps_values[j]))
  lines(grid, exact, col = "orange", lwd = 2)
}
```

Compare the tolerances with Q–Q plots.

```{r tolerance-qq, fig.width = 10, fig.height = 3.5}
par(mfrow = c(1, 3))
for (pair in list(c(1, 2), c(1, 3), c(2, 3))) {
  qqplot(draws[[pair[1]]], draws[[pair[2]]],
         xlab = paste("epsilon =", eps_values[pair[1]]),
         ylab = paste("epsilon =", eps_values[pair[2]]))
  abline(0, 1, lty = 2)
}
```

# What if we have twenty measurements?

A direct extension retains a dataset only when every simulated measurement is close to its observed counterpart: $\max_i |X_i-x_{\mathrm{obs},i}|<\epsilon$.

```{r twenty-measurements}
set.seed(681)
x_obs <- rnorm(20, 0.8, 1)   # observed data
X <- rnorm(20, 0.8, 1)       # simulated at theta = 0.8

plot_match <- function(x_obs, X, eps) {
  near <- abs(X - x_obs) < eps
  i <- seq_along(x_obs)
  plot(i, x_obs, pch = 4, ylim = range(x_obs - eps, x_obs + eps, X),
       xlab = "measurement i", ylab = "value",
       main = paste(sum(near), "of", length(X), "within", eps))
  segments(i, x_obs - eps, i, x_obs + eps)
  points(i, X, pch = 16, col = ifelse(near, "darkgreen", "gray"))
}
plot_match(x_obs, X, eps = 0.5)
```

Each simulated measurement has to land inside its own bar.

# How might we check that the sample mean is sufficient?

We will compare two summaries of the twenty measurements: $S(X)$ = mean of all twenty (sufficient) and $S(X)$ = mean of the first ten (not sufficient).

Row $i$ of `X` is twenty measurements simulated with $\theta_i$. Table rows: $\theta$ above its median; columns: $a^\top X$ above its median.

```{r check-summary}
set.seed(682)
n <- 20; N <- 200000
theta <- runif(N, -3, 3)
X <- matrix(rnorm(N * n, mean = theta), nrow = N)
check_summary <- function(S_X, S_x_obs, width = 0.05) {
  keep <- abs(S_X - S_x_obs) < width   # S(X) near S(x_obs)
  a <- rnorm(n)                      # random direction
  proj <- X[keep, ] %*% a            # projection a'X
  th <- theta[keep]
  table(th > median(th), proj > median(proj)) / sum(keep)
}

check_summary(rowMeans(X), mean(x_obs))
check_summary(rowMeans(X[, 1:10]), mean(x_obs[1:10]))
```

Four random directions: plot $\theta$ against $a^\top X$ for the kept runs, using the same four directions for both summaries. Orange lines are the medians; the numbers are the fractions of kept runs in each box.

```{r four-directions, fig.width = 12, fig.height = 6}
A <- matrix(rnorm(n * 4), nrow = n)   # four random directions
plot_directions <- function(S_X, S_x_obs, A, width = 0.05) {
  keep <- abs(S_X - S_x_obs) < width
  th <- theta[keep]
  for (k in 1:ncol(A)) {
    proj <- drop(X[keep, ] %*% A[, k])
    plot(proj, th, pch = 16, cex = 0.3, col = "gray40",
         xlab = paste0("a'X, direction ", k), ylab = "theta")
    abline(v = median(proj), h = median(th), col = "orange", lwd = 2)
    hi_t <- th > median(th); hi_p <- proj > median(proj)
    legend("topleft", sprintf("%.2f", mean(!hi_p & hi_t)), bty = "n")
    legend("topright", sprintf("%.2f", mean(hi_p & hi_t)), bty = "n")
    legend("bottomleft", sprintf("%.2f", mean(!hi_p & !hi_t)), bty = "n")
    legend("bottomright", sprintf("%.2f", mean(hi_p & !hi_t)), bty = "n")
  }
}

par(mfrow = c(2, 4))
plot_directions(rowMeans(X), mean(x_obs), A)                # top: mean of all 20
plot_directions(rowMeans(X[, 1:10]), mean(x_obs[1:10]), A)  # bottom: mean of first 10
```

Top: every box holds 1/4 of the runs, in every direction. Bottom: the boxes are uneven by different amounts in different directions.

Many random directions: correlations between $\theta$ and $a^\top X$ across 200 random directions. Dashed lines mark $\pm 2/\sqrt{\text{number of runs}}$.

```{r many-directions, fig.width = 9, fig.height = 3.5}
direction_cors <- function(S_X, S_x_obs, width = 0.05, K = 200) {
  keep <- abs(S_X - S_x_obs) < width
  A <- matrix(rnorm(n * K), nrow = n)   # K random directions
  drop(cor(theta[keep], X[keep, ] %*% A))
}
summaries <- list("Mean of all 20" = list(rowMeans(X), mean(x_obs)),
                  "Mean of first 10" = list(rowMeans(X[, 1:10]), mean(x_obs[1:10])))

par(mfrow = c(1, 2))
for (nm in names(summaries)) {
  S_X <- summaries[[nm]][[1]]; S_x_obs <- summaries[[nm]][[2]]
  r <- direction_cors(S_X, S_x_obs)
  hist(r, breaks = seq(-0.6, 0.6, by = 0.025), main = nm,
       xlab = "correlation of theta with a'X")
  abline(v = c(-2, 2) / sqrt(sum(abs(S_X - S_x_obs) < 0.05)), lty = 2)
}
```

# A learned summary for the rain model

Candidate features: rainy-day count, weather changes, endpoint weather, and longest rainy run. Use simulations to learn how these features jointly predict $\theta$.

```{r rain-features}
rain_features <- function(N, days = 6) {
  theta <- runif(N)
  X <- matrix(0L, N, days)
  X[, 1] <- rbinom(N, 1, theta)
  for (d in 2:days) {
    redraw <- runif(N) >= 0.5           # otherwise weather persists
    X[, d] <- X[, d - 1]
    X[redraw, d] <- rbinom(sum(redraw), 1, theta[redraw])
  }
  longest <- current <- integer(N)
  for (d in 1:days) {
    current <- ifelse(X[, d] == 1, current + 1L, 0L)
    longest <- pmax(longest, current)
  }
  data.frame(theta, count = rowSums(X),
             changes = rowSums(X[, -1] != X[, -days]),
             first = X[, 1], last = X[, days], longest)
}

set.seed(687)
train <- rain_features(150000)
fit <- lm(theta ~ count + changes + first + last + longest, data = train)
round(coef(fit), 3)
```

The learned summary $S(X)$ is the fitted prediction. For DRRRRD:

```{r rain-observed}
obs <- data.frame(count = 4, changes = 2, first = 0, last = 0, longest = 4)
predict(fit, obs)
```

The learned summary on held-out simulations.

```{r rain-heldout, fig.width = 5, fig.height = 5}
set.seed(688)
heldout <- rain_features(50000)
pred <- predict(fit, heldout)
show <- sample(nrow(heldout), 5000)
plot(pred[show], heldout$theta[show], pch = 16, cex = 0.3,
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Learned summary / prediction", ylab = "Generating theta")
abline(0, 1, lty = 2)
```

# Regression adjustment

Twenty measurements summarized by their mean, with $S(x_{\text{obs}}) = 0.8$ and a wide tolerance of 0.5. The mean of twenty $N(\theta, 1)$ measurements is $N(\theta, 1/20)$, so we simulate it directly.

```{r adjustment-bank}
set.seed(690)
N <- 1000000
theta <- runif(N, -3, 3)
S_X <- rnorm(N, theta, 1 / sqrt(20))

eps <- 0.5
keep <- abs(S_X - 0.8) < eps
theta_kept <- theta[keep]
d <- S_X[keep] - 0.8                    # d_i = S(X_i) - S(x_obs)
w <- 1 - (d / eps)^2                    # closer summaries get more weight
```

Fit $\theta_i=\alpha+\beta\,d_i+e_i$ to the kept draws and use $\theta_i^{\mathrm{adj}}=\widehat\alpha+\widehat e_i = \theta_i - \widehat\beta\, d_i$.

```{r adjustment-fit}
local_fit <- lm(theta_kept ~ d, weights = w)
theta_adj <- theta_kept - coef(local_fit)["d"] * d
```

Compare with tolerance 0.05, unadjusted. The orange curve is the exact posterior.

```{r adjustment-compare, fig.width = 10, fig.height = 3.5}
theta_narrow <- theta[abs(S_X - 0.8) < 0.05]
exact <- dnorm(grid, 0.8, 1 / sqrt(20))

par(mfrow = c(1, 3))
panels <- list("Tolerance 0.5, unadjusted" = list(theta_kept, rep(1, length(theta_kept))),
               "Tolerance 0.5, adjusted" = list(theta_adj, w),
               "Tolerance 0.05, unadjusted" = list(theta_narrow, rep(1, length(theta_narrow))))
for (nm in names(panels)) {
  v <- panels[[nm]][[1]]; wt <- panels[[nm]][[2]]
  breaks <- seq(-1, 2.6, by = 0.05)
  dens <- tapply(wt, cut(v, breaks), sum) / (sum(wt) * 0.05)   # weighted histogram
  plot(breaks[-1] - 0.025, ifelse(is.na(dens), 0, dens), type = "h", lwd = 4,
       col = "steelblue", xlim = c(-0.6, 2.2), ylim = c(0, 2),
       xlab = "theta", ylab = "Density", main = nm)
  lines(grid, exact, col = "orange", lwd = 2)
}

wsd <- function(v, wt) sqrt(sum(wt * (v - weighted.mean(v, wt))^2) / sum(wt))
c(unadjusted = sd(theta_kept), adjusted = wsd(theta_adj, w), narrow = sd(theta_narrow))
```
