## Avoiding iteration, split/apply/combine

Reading: R Cookbook, Chapter 6

Agenda for today:

- Homework 1 solutions

- Avoiding iteration in base R

- Split/apply/combine

-----

## Avoiding iteration

We've already seen a lot of examples of functions in R that are written so as to avoid iteration.


Built-in functions for making sequences or repeating elements, for example, we can say:

```{r}
x = seq(1, 11, by = 2)
```

instead of

```{r}
x = numeric(6)
for(i in 1:length(x)) {
    x[i] = 1 + 2 * (i - 1)
}
```

-----

All the vectorized functions, so we can use

```{r}
log_x = log(x)
```

instead of

```{r}
log_x = numeric(length(x))
for(i in 1:length(x)) {
    log_x[i] = log(x[i])
}
```

-----

Plus a whole bunch we haven't seen yet, like `rowSums`, `colMeans`, `sweep`, `scale`

However, not everything is vectorized, and you will often want to go beyond the built-in functions, and for that you use the `*apply` family of functions.

## Applying to one-dimensional structures

`lapply`, and `sapply` will apply a function to a one-dimensional structure.

The difference between them is how the output is formatted.

They have the same syntax: `lapply(x, fun, ...)`, `sapply(x, fun, ...)`

- `x` is a vector or list that the function should be applied to

- `fun` is the function that should be applied to each element of the vector or the list

- You can specify other arguments in the `...` portion. `fun` will be called with first argument equal to each element of `x`, and subsequent arguments anything you specify in the `...` portion.

. . .

Return values:

- `lapply` returns a list with length equal to the length of `x` containing the result of the function evaluation on the corresponding element of `x`.

- `sapply` tries to simplify the output to a vector or matrix. If each element of the output is of length 1 it returns a vector. If each element has the same length of two or more, it returns a matrix. Otherwise it returns a list like `lapply`.

-----

Examples from the homework:

. . .

```{r}
library(readr)
book_dir = "~/GitHub/stat-comp-spring-19/homeworks/hw1/books"
book_file_names = list.files(book_dir, full.names = TRUE)
books = lapply(book_file_names, read_file)
```

We need to do this because `read_file` doesn't vectorize. Check what happens if you do `read_file(book_file_names)` (you'll have to change `book_dir` to wherever you saved the books to).

-----

 . . .

Example using extra function arguments:

```{r}
get_num_characters = function(file_name, honorific_regex) {
    text = read_file(file_name)
    name_matches = regmatches(text, gregexpr(honorific_regex, text))[[1]]
    return(length(unique(name_matches)))
}
first_five_books = list.files(book_dir, full.names = TRUE)[1:5]
lapply(first_five_books, get_num_characters, "(((Mr|Ms|Mrs|Dr)\\.?)|Miss) ([A-Z]([a-z]+|\\.)[ ]+)+[A-Z][a-z]+")
sapply(first_five_books, get_num_characters, "(((Mr|Ms|Mrs|Dr)\\.?)|Miss) ([A-Z]([a-z]+|\\.)[ ]+)+[A-Z][a-z]+")
```


## Applying to two-dimensional structures

Syntax: `apply(mat, dim, fun, ...)`

- `mat`: The matrix or data frame you want to apply the function to.

- `dim`: Either 1 or 2: 1 means apply the function to each row in `mat`, 2 means apply the function to each column in `mat`. (1 and 2 are related to the way the dimensions are laid out: the first dimension in the matrix is the number of rows, the second is the number of columns.)

- `fun`: The function you want to apply to each row or to each column. It should take a vector as its first argument.

- `...`: Extra arguments to be passed to `fun`.

As with `lapply` and `sapply`, `fun` will be called with each row or each column of `mat` as its first argument, along with any extra arguments specified after the function.

-----

Some simple examples:

. . .

```{r, echo=FALSE}
set.seed(0)
```
```{r}
X = matrix(sample(-5:5, size = 20, replace = TRUE), nrow = 4, ncol = 5)
X
apply(X, 2, min)
```
----

. . .

```{r}
log_X = log(X)
log_X
apply(log_X, 1, min, na.rm = TRUE)
```

-----

Be careful when using apply on rows of a data frame: apply assumes the argument you pass to the function is a vector, and will coerce the rows of the data frame to vectors to make it so.

```{r}
steak_combinations = data.frame(
    temp = c(116, 120, 135, 105),
    type = c("rare", "med_rare", "med_rare", "rare"))

steak_directions_for_apply = function(temp_and_time) {
    temp = temp_and_time[1]
    steak_type = temp_and_time[2]
    if(steak_type == "rare" & temp > 115) {
        return("take your steak off!")
    } else if(steak_type == "med_rare" & temp > 125) {
        return("take your steak off!")        
    } 
    "you can keep cooking"
}
## this works, but it really shouldn't
apply(steak_combinations, 1, steak_directions_for_apply)
```

-----

If you want to apply a function that uses columns of a data frame that are of different types, better/safer to use `mapply`:

`mapply(f, vec1, vec2, ..., vecN)`

- `f` is the function you want to apply

- Output is a list, ith element will be `f(vec1[i], vec2[i], ..., vecN[i])`

-----

```{r}
steak_directions = function(steak_type, temp) {
    if(steak_type == "rare" & temp > 115) {
        return("take your steak off!")
    } else if(steak_type == "med_rare" & temp > 125) {
        return("take your steak off!")        
    } 
    "you can keep cooking"
}

mapply(FUN = steak_directions, steak_combinations$type, steak_combinations$temp)
```


## What if things aren't quite so tidy?

- The `*apply` functions we've seen so far have applied a function element-by-element to a vector or matrix.

- What if we need to apply a function not to individual elements/rows/columns, but to subsets of different sizes?

## Split/apply/combine paradigm

Turns out this is common enough that it has a name and a bunch of built-in functions.

The programming pattern is pretty clear from the name, what we want to do is:

- split the data into groups

- apply a function to each of the groups

- combine the results of applying the function to each group into an output dataset

## Why is this a useful abstraction?

Imagine how you would do this if you only had for-loops.

You would have to first

- Compute what size the output dataset should be

- Create a data structure to hold the output.

Then, for each level of the grouping factor:

- Compute how many data elements correspond to that grouping level

- Create a data structure to hold the data subset

- Loop through the data to extract the correct subset

- Apply your function to that subset and put the output in the data structure your created at the beginning.

-----

Compare with split/apply/combine abstraction:

- Specify data

- Specify grouping variable

- Specify function to apply

The abstraction is useful both for writing and reading the resulting code: you think about _what_ you want to do instead of _how_ to do it.

## Split/apply/combine for vectors

Simple case: we have a vector, and a grouping factor, we want to apply a function to groups of elements of the vector defined by the grouping factor

Solution is tapply

`tapply(x, f, fun)`

Syntax:

- `x`: A vector with the data.

- `f`: A factor variable describing how the data should be split up.

- `fun`: The function you want to apply to each subset of the data.

-----

Example: data on police stops in New York City in 1998--1999, information on the number of stops as categorized by certain variables.

In particular, we have counts of police stops for all combinations of

- precinct: 75 total

- eth: Ethnicity of the person stopped, three possibilities (1 = black, 2 = Hispanic, 3 = white), and

- crime: The type of crime, four possibilities (1 = violent, 2 = weapons, 3 = property, and 4 = drug)

```{r}
frisk =
    read.table("http://www.stat.columbia.edu/~gelman/arm/examples/police/frisk_with_noise.dat",
               skip = 6, header = TRUE)
head(frisk)
```

-----

Suppose we want to get the total number of stops in each precinct.

We can use `tapply` with the `sum` function to do so:

```{r}
tapply(frisk$stops, frisk$precinct, sum)
```

Just as a check, make sure that the numbers match up:
```{r}
precinct_1 = subset(frisk, precinct == 1)
sum(precinct_1$stops)
```

-----

And remember how much better this is than iterating using a for loop would have been:
```{r}
## find the number of precincts and make a vector of the right length to store the output
precincts = unique(frisk$precinct)
stops_per_precinct = numeric(length(precincts))
## iterate over precinct, get the subset for each one, compute the number of stops, and store
for(i in 1:length(precincts)) {
    precinct_subset = subset(frisk, precinct == precincts[i])
    precinct_stops = sum(precinct_subset$stops)
    stops_per_precinct[i] = precinct_stops
}
stops_per_precinct
```


-----

## Split/apply/combine for data frames

Splitting a data frame into groups of rows and applying a function to the groups:

`by(df, fact, fun)`

- `df`: A data frame containing the data you want to split up and apply a function to.

- `fact`: A factor variable describing the groups you want to split `df` into.

- `fun`: The function you want to apply to each group. It should take as its first argument a data frame.

------

Example:

```{r}
data(diamonds)
get_diamond_coefficients = function(data_subset) {
    diamond_lm = lm(log(price) ~ carat, data = data_subset)
    diamond_coefficients = coef(diamond_lm)
    return(diamond_coefficients)
}
## by does the split and apply steps, and an ugly version of a combine step
out = by(diamonds, diamonds$color, get_diamond_coefficients)
```

-----

The output is as a list with some extra attributes (the class is `by`, as you can check).

The form is not ideal on its own, usually you'll want to simplify it.
You have a couple of options, including the examples below:

```{r}
simplify2array(out)
## changing the output from a list to a matrix
do.call(rbind, out)
## it's kind of bad that the splitting variable is only available as row names
## we can fix that but it's a bit of a pain
diamond_coef_matrix = do.call(rbind, out)
diamond_coef_df = data.frame(diamond_coef_matrix,
    color = attributes(out)$dimnames$`diamonds$color`)
diamond_coef_df
```

-----

`by` can also be used with multiple grouping factors:

```{r}
by(diamonds, diamonds[, c("color", "cut")], get_diamond_coefficients)
```

But how do you get at which coefficients correspond to which factor groups?

-----


## If you want to do split and apply/combine separately

`split` does just the splitting part of split/apply/combine

Syntax: `split(x, f)`

- `x`: Your data, either a vector or a data frame, that you want to divide into groups.

- `f`: A factor variable used to split `x`. Length is equal to the length of `x` if it is a vector or equal to the number of rows of `x` if it is a data frame.

Output:

- A list of length equal to the number of levels in `f`

- Each element of the list is either a vector or a data frame, depending on what `x` was, containing the elements of `x` corresponding to one of the levels of `f`.


-----

We can re-do the stop-and-frisk example that we used `tapply` for with `split` followed by `lapply` or `sapply`

```{r}
frisk_split_by_precinct = split(frisk$stops, frisk$precinct)
## check that this has length 75, the number of precincts
length(frisk_split_by_precinct)
sapply(frisk_split_by_precinct, sum)
```

-----

### Anonymous functions

If you have a function you only want to use once, you don't have to assign it to anything.

When you use a function this way it is called an _anonymous function_ because it doesn't have a name in your code.

These are often used in the context of the apply family.

-----

```{r}
diamonds_split = split(diamonds, diamonds$color)
## here the function for computing the coefficients is an anonymous function
diamond_coef_list = lapply(diamonds_split, FUN = function(data_subset) {
    diamond_lm = lm(log(price) ~ carat, data = data_subset)
    diamond_coefficients = coef(diamond_lm)
    return(diamond_coefficients)
})
## changing the output from a list to a matrix
do.call(rbind, diamond_coef_list)
```

-----

In general, base R apply-family functions don't produce nice output from the combine step.

We'll see implementations of split/apply/combine next time that do a nicer job.
