## Testing and debugging

Logistics

- I'll post a study guide for the midterm on Sunday

Today:

- Finish debugging

- Testing, test-based design


Reading

- [Wickham, "testthat: Get Started with Testing"](https://journal.r-project.org/archive/2011/RJ-2011-002/RJ-2011-002.pdf). The details are slightly out of date compared with the current version of the package, but the philosophy is the same.


## Informal testing

- We've done informal testing when debugging, and you've probably done it on your own.

- Check whether the output is what we expect, either by inspection or using `==` or `identical` or something similar.



## Simple cases

Idea: make sure your function works on cases where you know what the answer should be.

You're checking that the core behavior is correct.

-----



```{r}
## returns the minimum value of d[i,j], i != j, and
## the row/col attaining that minimum, for square
## symmetric matrix d; no special policy on ties;
## motivated by distance matrices
mind = function(d) {
    n = nrow(d)
    ## add a column to identify row number for apply()
    dd = cbind(d, 1:n)
    wmins = apply(dd[-n, ], 1, imin)
    ## wmins will be 2xn, 1st row being indices and 2nd being values
    i = which.min(wmins[1, ])
    j = wmins[2, i]
    return(c(d[i, j], i, j))
}

## finds the location, value of the minimum in a row x
imin = function(x) {
    n = length(x)
    i = x[n]
    j = which.min(x[(i + 1):(n - 1)])
    return(c(j, x[j]))
}

m = rbind(c(0, 12, 5), c(12, 0, 8), c(5, 8, 0))
```

-----

. . .

Let's write some tests for `imin`.

The comment says that it finds the location and value of the minimum in a row x, so let's see if it does.

```{r}
## location of the minimum
x = 1:5
index_and_value = imin(x)
## should be 1, per the comment
index_and_value[1]
## should be 1, per the comment
index_and_value[2]
```

-----

The comment was misleading: what the function is supposed to do is to take a vector x whose last element indicates the row the vector was taken from, and finds the minimum index among locations corresponding to the upper triangle of the initial matrix.

This is very confusing, and it's why there was a bug in the function to begin with.

We might be tempted to change the function in the following way:

```{r}
imin = function(x) {
    n = length(x)
    row = x[n]
    upper_tri_idx = (row + 1):(n - 1)
    idx_in_upper_tri = which.min(x[upper_tri_idx])
    idx = upper_tri_idx[idx_in_upper_tri]
    value = x[idx]
    return(c(idx, value))
}
```

-----

. . .

And then we can test:

```{r}
row = 1
x = c(5:1, row)
index_and_value = imin(x)
## index of the minimum should be 5
index_and_value[1] == 5
## value of the minimum should be 1
index_and_value[2] == 1
```

## Edge cases

Idea: If the input to the function isn't exactly what you expect, what happens?

- You're checking that if something funny happens (bad input from a user or another function), your function will (best case) still work correctly or (at minimum) fail informatively.

- Very important to make sure that the function doesn't fail silently: it looks like it's producing good results, but they're actually wrong.

- Most important if you are not directly providing the input to the function.

-----

. . .

Let's try testing the `imin` function again with an edge case:

```{r}
row = 5
x = c(5:1, row)
index_and_value = imin(x)
## there aren't any elements in the upper triangle in row 5, so these should be some sort of NA
index_and_value
```

-----

Based on our test, we might modify our function to look like this:

```{r}
imin = function(x) {
    n = length(x)
    row = x[n]
    if(row >= (length(x) - 1)) {
        upper_tri_idx = c()
    } else {
        upper_tri_idx = (row + 1):(n - 1)
    }
    idx_in_upper_tri = which.min(x[upper_tri_idx])
    idx = upper_tri_idx[idx_in_upper_tri]
    value = x[idx]
    return(c(idx, value))
}
```

-----

. . .

And try testing again:

```{r}
row = 5
x = c(5:1, row)
index_and_value = imin(x)
## there aren't any elements in the upper triangle in row 5, so these should be some sort of NA
index_and_value
```

## The real moral of the story

This is a bad way to write functions to compute the minimum, we should throw it all out and start over.

## Test-based design

- Decide what you want your function(s) to do

- Write tests for those behaviors

- Write the functions, check whether they pass the tests

- If they pass, you're done! Otherwise, cycle through changing the functions and testing.

## Example

I want to make a program that performs gradient descent for functions where I don't have the derivative in closed form.

I'm going to need to write a function that computes the derivative


```{r, eval = FALSE}
## derivative of x^2, evaluated at x = 1
deriv(function(x) x^2, 1) == 2
## derivative of 2 * x, evaluated at x = -5
deriv(function(x) 2 * x, -5) == 2
## derivative of x^2, evaluated at x = 0
deriv(function(x) x^2, 0) == 0
## derivative of e^x, evaluated at x = 0
deriv(function(x) exp(x), 0) == exp(0)
```

------

Then I write the following function, based on advice from wikipedia:

```{r}
deriv = function(fn, x) {
    eps = .Machine$double.eps
    h = sqrt(eps) * x
    deriv = (fn(x + h) - fn(x - h)) / (2 * h)
    return(deriv)
}
```

-----

. . .

And run through my tests:

```{r}
## derivative of x^2, evaluated at x = 1
deriv(function(x) x^2, 1) == 2
## derivative of 2 * x, evaluated at x = -5
deriv(function(x) 2 * x, -5) == 2
## derivative of x^2, evaluated at x = 0
deriv(function(x) x^2, 0) == 0
## derivative of 2 * x, evaluated at x = 0
deriv(function(x) 2 * x, 0) == 2
```

The third and fourth tests failed, and not just because of precision. Why?

------

Then we can modify the function to evaluate derivatives at $x = 0$

```{r}
deriv = function(fn, x) {
    eps = .Machine$double.eps
    if(x == 0) {
        h = 2 * eps
    } else {
        h = sqrt(eps) * x
    }
    deriv = (fn(x + h) - fn(x - h)) / (2 * h)
    return(deriv)
}
```

------

. . .

Run through the tests again:

```{r}
## derivative of x^2, evaluated at x = 1
deriv(function(x) x^2, 1) == 2
## derivative of 2 * x, evaluated at x = -5
deriv(function(x) 2 * x, -5) == 2
## derivative of x^2, evaluated at x = 0
deriv(function(x) x^2, 0) == 0
## derivative of 2 * x, evaluated at x = 0
deriv(function(x) 2 * x, 0) == 2
## derivative of e^x, evaluated at x = 0
deriv(function(x) exp(x), 0) == exp(0)
```


## More formal ways of integrating tests

- R package called `testthat`

- Aimed more at package developers

- Allows for tests to be stored and run automatically.


-----

Suppose we have `testthat_example.R` and `numerical_deriv.R`, with contents that look like this:


`testthat_example.R`:

```{r, eval = FALSE}
context("Check numerical derivative")
source("numerical_deriv.R")

test_that("derivatives match on simple functions", {
    expect_equal(deriv(function(x) x^2, 1), 2)
    expect_equal(deriv(function(x) 2 * x, -5), 2)
    expect_equal(deriv(function(x) x^2, 0), 0)
    expect_equal(deriv(function(x) 2 * x, 0), 2)
    expect_equal(deriv(function(x) exp(x), 0), exp(0))
})

test_that("error thrown when derivative doesn't exist", {
    expect_error(deriv(function(x) log(x), 0))
}) 
```

`numerical_deriv.R`:

```{r, eval = FALSE}
deriv = function(fn, x) {
    eps = .Machine$double.eps
    if(x == 0) {
        h = 2 * eps
    } else {
        h = sqrt(eps) * x
    }
    deriv = (fn(x + h) - fn(x - h)) / (2 * h)
    return(deriv)
}
```

-----

```{r}
library(testthat)
test_dir(".")
```


## Idea behind testthat

- Expectations: Finest unit of testing, checks one aspect of a function's output.

- Tests: Groups of related expectations.

- Contexts/files: Each context or file can contain a group of tests. Primarily useful for having the test output formatted nicely. Perhaps also useful if you have some tests that require a lot of setup and you don't want to run them every time.

## Expectations

An expectation is the finest unit of testing, tests whether a call to a function does what you expect.

- All start with `expect_`

- All have two arguments: the actual result, and what you expect

- `testthat` will throw an error if the two don't match

-----

. . .

Some of the most useful expectations

- `expect_equal`/`expect_identical`: Check for equality within numerical precision or exact equivalence (`expect_identical` built on `identical` function, which also checks for type)

. . .

- `expect_match`: Checks whether a string matches a regular expression

- `expect_output`: Checks the output of a function the same way `expect_match` would
```{r}
a = list(1:10, letters)
str(a)
expect_output(str(a), "List of 2")
expect_output(str(a), "int [1:10]", fixed = TRUE)
```

- `expect_warning`/`expect_error`: Checks whether the function gives an error or a warning.

. . .

- `expect_is`: Checks whether the function gives a result of the correct class (we'll talk more about classes in a couple weeks).

. . .

- `expect_true`/`expect_false`: Catch-alls for cases the other expectations don't cover.


## Tests

- For `testthat`, a test is just a group of expectations.

- You can group them however you like, but usually you think of them as covering one unit of functionality. Often this means one test per function.

- Group expectations into tests so that when a test fails, it's easy to figure out what part of the code caused the error.


## Advice on writing tests

- When you add a new function/functionality, add a new test.

- Write a test when you discover a bug. 

- Most important to test code that is delicate: has complicated dependencies on other functions many edge cases, doing something complicated that you're not sure about. (In that case you might want to re-think your function design though.)

- If tests are grouped according to the desired behavior of a function, they are easier to update later if you want to change the behavior of the function.
