## Object-oriented programming


```{r setup, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(fig.cap="",
               fig.path="lecture-13-fig/")
```


Reading: Matloff, Chapter 9

Agenda for today:

- Objects and classes in R

- Continue the bootstrap example

-----

## What is object-oriented programming?

Two main concepts, _classes_ and _methods_.

- A class defines a type of object: what sort of data is associated with it, what are valid values for the data, what methods can be called on it.

- Every object is an instance of a class.

- A method is a function associated with a particular type of object.


## In R

- S3: Very informal versions of classes, lists with a "class" attribute, allowing for one-argument method dispatch.

- S4: Formal classes, class definitions, methods, considered safer but with more overhead.

- Both types support _generic functions_: functions that have different behavior depending on the class of the arguments passed to them.

## S3 classes

- An S3 class is a list with an extra class attribute.

- Class instantiation is by creating a list and setting the class attribute.

- Object orientation is only through generic functions.

-----

For example:

```{r}
joe = list(name = "Joe", salary = 55000, union = TRUE)
class(joe) = "employee"
attributes(joe)
print(joe)
```

## Generic functions/method dispatch for S3 classes

- A _generic function_ is a function whose behavior depends on what class its arguments are.

- We have seen functions like `print`, `plot`, `summary`, all of which are generic.

- Generic functions are very simple: usually just a call to `UseMethod`

```{r}
print
plot
mean
```

------

Methods:

- A _method_ is the function that is actually called on a specific class.

- In S3, a method for a given class and generic function is just a function with name `generic.class`.

- For example: `print.lm`, `plot.lm` are the methods for the `lm` class associated with the `print` and `plot` generic functions.

- To see that they are really just normal functions, try typing `stats:::plot.lm` or `stats:::print.lm` in your R console.

- To define a method associated with a class you have created,  define a function with name `generic.class`

. . .

Method dispatch:

- Method dispatch refers to how R decides what method to use when a
generic functions is called.

- The `UseMethod` function is what does this for S3, and it just works
by name matching.

-----

Example:

```{r}
print.employee = function(x) {
    cat(x$name, "\n")
    cat("salary", x$salary, "\n")
    cat("union member", x$union, "\n")
}
print(joe)
joe
```

## S4 classes


S4 classes have three properties:

- A class _name_

- A _representation_: A list of _slots_ giving names and classes for the objects associated with the class.

- A vector of classes it inherits from (we won't use inheritance in this class, but you should know it's available to you).


## S4 class definition

Syntax: `setClass(class_name, class_representation, contains)`

- `class_name` is the name of the class.

- `class_representation` is a list decribing the slots and their types.

- `contains` describes the inheritance.

For example:

```{r, eval = FALSE}
setClass("employee",
         representation(
             name = "character",
             salary = "numeric",
             union = "logical"))
```

------

Note: This function breaks one of our rules from the beginning: it's called for its side effect. It assigns an object defining the class, and also returns invisibly a class generation function.


```{r}
rm(list = ls(all.names = TRUE))
ls(all.names = TRUE)
setClass("employee",
         representation(
             name = "character",
             salary = "numeric",
             union = "logical"))
ls(all.names = TRUE)
.__C__employee
```

------

Don't use `setClass` this way: it's just to show you that the method returns a class creation function.

```{r}
class_creation_fn = setClass("employee",
         representation(
             name = "character",
             salary = "numeric",
             union = "logical"))
jane = class_creation_fn(name = "Jane", salary = 55000, union = FALSE)
jane
```

## S4 class instantiation

- To make an object of a given S4 class, use `new`

- Syntax: `new(class, representation)`

```{r}
joe = new("employee", name = "Joe", salary = 55000, union = TRUE)
joe
```

## Data access in S4 classes

Slot access is with `@`, not `$`: `object@slot` will give the data associated with `slot` in `object`.

```{r}
joe@salary
joe$salary
```

## Generic functions and methods for S4 classes

Remember:

- A generic function is a function whose behavior depends on the class of its arguments.

- A method is the function associated with a specific combination of generic function and argument classes.

. . .

Syntax for setting a method associated with a generic function: `setMethod(generic, signature, fn)`

- `generic` is a string specifying the generic function for which we want to specify a class-specific method.

- `signature` describes the classes of the arguments.

- `fn` is the function we want to use for tha specified combination of generic function and argument classes.


------

. . .

For example: `show` is a generic function used to print S4 objects.

We can create a method associated with the `show` generic function and the `employee` S4 class as follows:

```{r}
setMethod("show", signature = signature("employee"), definition = function(object) {
    inorout = ifelse(object@union, "is", "is not")
    cat(object@name, "has a salary of", object@salary, "and", inorout, "in the union", "\n")
})
show(joe)
joe
```

## A longer example

Remember our bootstrap example from last time?

Now, in addition to computing confidence intervals, we want to plot the bootstrap sampling distributions.

Last time we settled on the following set of functions.

```{r}
bootstrap_ci = function(data, estimator, alpha, B) {
    boot_estimates = get_boot_estimates(data, estimator, B)
    boot_ci = get_ci(boot_estimates, alpha)
    return(boot_ci)
}

get_boot_estimates = function(data, estimator, B) {
    boot_estimates = replicate(B, expr = {
        boot_data = get_bootstrap_sample(data)
        boot_estimate = estimator(boot_data)
        return(boot_estimate)
    })
    return(boot_estimates)
}

get_ci = function(estimates, alpha) {
    ci_lo = alpha / 2
    ci_hi = 1 - (alpha / 2)
    if(!is.null(dim(estimates))) {
        ## if we have multi-dimensional estimates
        cis = plyr::aaply(estimates, 1, function(x) quantile(x, probs = c(ci_lo, ci_hi)))
    } else {
        ## if we have one-dimensional estimates
        cis = quantile(estimates, probs = c(ci_lo, ci_hi))
    }
    return(cis)
}

get_bootstrap_sample = function(data) {
    if(!is.null(dim(data))) {
        ## in this case, data is rectangular, and we want to sample rows
        n = dim(data)[1]
        boot_idx = sample(1:n, size = n, replace = TRUE)
        bootstrap_sample = data[boot_idx,]
    } else {
        ## in this case, data is a vector and we want to sample elements of the vector
        n = length(data)
        boot_idx = sample(1:n, size = n, replace = TRUE)
        bootstrap_sample = data[boot_idx]
    }
    return(bootstrap_sample)
}
```

-----

Notice that the output from `get_boot_estimates` could be used for a lot of different tasks

- Last time we used it to get bootstrap confidence intervals

- We could also use it to get bootstrap standard errors, or make plots, or do anything else that depends on the set of bootstrap samples.

- Let's try making it into its own class and setting some methods for common operations.


------

First step: Modify the function so it returns something with a `class` attribute.

```{r}
get_boot_estimates = function(data, estimator, B) {
    boot_estimates = replicate(B, expr = {
        boot_data = get_bootstrap_sample(data)
        boot_estimate = estimator(boot_data)
        return(boot_estimate)
    })
    class(boot_estimates) = "boot_dist"
    return(boot_estimates)
}
```

-----

Setond step: Create methods for the `boot_dist` class associated with the `plot` and `print` generic functions.

```{r}
library(ggplot2)
plot.boot_dist = function(x) {
    ggplot(data.frame(boot_samples = as.vector(x))) +
        geom_histogram(aes(x = boot_samples)) +
        ggtitle("Bootstrap distribution")
}
print.boot_dist = function(x) {
    n = length(x)
    cat("Bootstrap distribution object,", n, "bootstrap samples\n")
    cat("Bootstrap standard error:", sd(x), "\n")
}
```

-----

Check whether it works:

```{r}
boot_dist = get_boot_estimates(rnorm(1:10), estimator = mean, B = 10000)
boot_dist
plot(boot_dist)
```

## Example with S4 classes

The functions above only work if we're getting bootstrap distributions for one parameter at a time.

```{r}
data(iris)
iris_coef_estimator = function(d) {
    iris_lm = lm(Sepal.Length ~ Sepal.Width + Petal.Length, data = d)
    iris_coef = coef(iris_lm)
    return(iris_coef)
}
boot_dist = get_boot_estimates(iris, iris_coef_estimator, B = 1000)
```

For the next example, we'll both fix this problem and show how you would use S4 classes instead of S3 classes.

-----

First step: set an S4 class for the bootstrap distribution.

```{r}
setClass("boot_dist",
         representation = list(boot_samples = "matrix", nparams = "numeric", nboot = "numeric"))
ls(all.names = TRUE)
```

------

Then we modify the `get_boot_estimates` function to return an object of the `boot_dist` class.

```{r}
get_boot_estimates = function(data, estimator, B) {
    boot_estimates = replicate(B, expr = {
        boot_data = get_bootstrap_sample(data)
        boot_estimate = estimator(boot_data)
        return(boot_estimate)
    })
    boot_dist_object = make_bd_object(boot_estimates)
    return(boot_dist_object)
}
```

-----

```{r}
## takes either a vector or a matrix and creates a boot_dist object
make_bd_object <- function(estimates) {
    if(is.null(dim(estimates))) { ## if estimates is a vector
        nparams = 1
        nboot = length(estimates)
        estimates = matrix(estimates, nrow = 1)
    } else { ## if estimates is a matrix
        nparams = nrow(estimates)
        nboot = ncol(estimates)
    }
    bd = new("boot_dist", boot_samples = estimates, nparams = nparams, nboot = nboot)
    return(bd)
}
```

-----

Next step: set method corresponding to the `show` generic:

```{r}
setMethod("show", signature = "boot_dist", function(object) {
    cat("Bootstrap distribution object,", object@nboot, "bootstrap samples\n")
    cat("Number of parameters:", object@nparams, "\n")
    cat("Bootstrap estimate of standard error:", apply(object@boot_samples, 1, sd), "\n")
})
```

-----

Set method corresponding to the `plot` generic:

```{r}
setMethod("plot", signature = "boot_dist", function(x) {
    melted_samples = melt(x@boot_samples)
    if(x@nparams == 1) {
        ggplot(melted_samples) +
            geom_histogram(aes(x = value)) +
            ggtitle("Bootstrap distribution")
    } else {        
        ggplot(melted_samples) +
            geom_histogram(aes(x = value)) +
            facet_wrap(~ Var1, scales = "free") +
            ggtitle("Bootstrap distributions for each parameter")
    }
})
```

-----

And finally see whether it works:

```{r}
boot_dist = get_boot_estimates(rnorm(10), estimator = mean, B = 100)
boot_dist
plot(boot_dist)
```

-----

And for multiple parameters:

```{r, fig.width = 10, fig.height = 4}
boot_dist_iris_coef = get_boot_estimates(iris, iris_coef_estimator, B = 1000)
boot_dist_iris_coef
plot(boot_dist_iris_coef)
```
