% Stat 470/670 Lecture 12: More interactions, coplots, and modeling
% Julia Fukuyama

```{r setup, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(fig.cap="", fig.width = 5, fig.height = 3, dpi=175, fig.path="lecture-12-fig/", warning = FALSE)
load("../../datasets/lattice.RData")
library(magrittr)
library(tidyverse)
library(ggplot2)
library(broom)

make_coplot_df = function(data_frame, faceting_variable, number_bins = 6) {
    ## co.intervals gets the limits used for the conditioning intervals
    intervals = co.intervals(data_frame[[faceting_variable]], number = number_bins)
    ## indices is a list, with the ith element containing the indices of the
    ## observations falling into the ith interval
    indices = apply(intervals, 1, function(x)
        which(data_frame[[faceting_variable]] <= x[2] & data_frame[[faceting_variable]] >= x[1]))
    ## interval_descriptions is formatted like indices, but has interval
    ## names instead of indices of the samples falling in the index
    interval_descriptions = apply(intervals, 1, function(x) {
        num_in_interval = sum(data_frame[[faceting_variable]] <= x[2] & data_frame[[faceting_variable]] >= x[1])
        interval_description = sprintf("(%.2f, %.2f)", x[1], x[2])
        return(rep(interval_description, num_in_interval))
    })
    ## df_expanded has all the points we need for each interval, and the
    ## 'interval' column tells us which part of the coplot the point should
    ## be plotted in
    df_expanded = data_frame[as.vector(unlist(indices)),]
    df_expanded$interval = factor(unlist(interval_descriptions),
        levels = unique(unlist(interval_descriptions)), ordered = TRUE)
    return(df_expanded)
}
get_interval_mean = function(interval_description) {
    interval_description = gsub("\\(|\\)", "", interval_description)
    lo_and_hi = strsplit(interval_description, split = ", ")[[1]]
    interval_mean = mean(as.numeric(lo_and_hi))
    return(interval_mean)
}
get_interval_means = function(interval_descriptions){
    sapply(interval_descriptions, get_interval_mean)
}
```


## Today: building and checking models with trivariate data

One more example with a slightly complicated model

## Rubber data

Reading: Cleveland pp. 180-187, 200-213

The data frame rubber in `lattice.RData` contains three measurements on 30 specimens of tire rubber.

The variables are:

> - `hardness`: How much the rubber rebounds after being indented.

> - `tensile.strength`: The force per cross-sectional area required to break the rubber (in kg/cm2).

> - `abrasion.loss`: The amount of material lost to abrasion when rubbing it per unit energy (in grams per hp-hour). This gives you an idea how fast the tire will wear away when you drive. If we had to choose a “response” variable, it would be this one.


## Pairs plot of the three variables

```{r, fig.height=5}
library(GGally)
ggpairs(rubber[,c("hardness", "tensile.strength", "abrasion.loss")])
```

## Coplot of abrasion loss and tensile strength given hardness

```{r}
coplot_hardness = make_coplot_df(rubber, "hardness", number_bins = 8)
ggplot(coplot_hardness, aes(x = tensile.strength, y = abrasion.loss)) +
    geom_point() +
    facet_wrap(~ interval, ncol = 4) +
    stat_smooth(method = "lm", se = FALSE)
ggplot(coplot_hardness, aes(x = tensile.strength, y = abrasion.loss)) +
    geom_point() +
    facet_wrap(~ interval, ncol = 4) +
    stat_smooth(method = "loess", method.args = list(degree = 1), se = FALSE)
```

An easier way to make a coplot with ggplot is to use the `cut_number` function:
```{r}
ggplot(rubber, aes(x = tensile.strength, y = abrasion.loss)) +
    geom_point() +
    facet_grid(. ~ cut_number(hardness, n = 4)) +
    stat_smooth(method = "loess", method.args = list(degree = 1, span = .9), se = FALSE)
```


## Several versions of coplots of abrasion loss and hardness given tensile strength

```{r}
coplot_ts = make_coplot_df(rubber, "tensile.strength", number_bins = 8)
ggplot(coplot_ts, aes(x = hardness, y = abrasion.loss)) +
    geom_point() +
    facet_wrap(~ interval, ncol = 4) +
    stat_smooth(method = "lm", se = FALSE)
```


```{r}
ggplot(coplot_ts, aes(x = hardness, y = abrasion.loss)) +
    geom_point() +
    facet_wrap(~ interval, ncol = 4) +
    stat_smooth(method = "loess", method.args = list(span = .5, degree = 1), se = FALSE)
ggplot(coplot_ts, aes(x = hardness, y = abrasion.loss, color = interval)) +
    stat_smooth(method = "loess", method.args = list(span = .5, degree = 1), se = FALSE)
ggplot(coplot_ts, aes(x = hardness, y = abrasion.loss, color = interval)) +
    stat_smooth(method = "lm", se = FALSE)
```


## Building a model


Let's start off building a model with no interaction but with a non-linear function of `tensile.strength`.

To do this, we need to:

> - Decide on a non-linear function to use.

> - Implement this function in R.

> - Apply the function to `tensile.strength`.

## Deciding on a function

We want our non-linear function to be linear for values of `tensile.strength` below 180, flat for values above 180, and continuous. One such function is
$$
f(x) = \begin{cases}
x - 180 & x \le 180\\
0 & x > 180
\end{cases}
$$

## Writing the function in R

The way we would write this in R would be
```{r}
tslow = function(x) {
    return((x - 180) * (x < 180))
}
```

## Applying the function to `tensile.strength`

And to create a variable corresponding to this transformation of tensile strength, we could use

```{r, eval = FALSE}
rubber %>% mutate(ts.low = tslow(tensile.strength))
```

However, we don't need to do that because Cleveland has already done it for us (the variable `ts.low` already exists in the data set and is exactly this function of `tensile.strength`).

## Fitting and visualizing the model

To fit the model:

```{r}
library(MASS)
rubber.rlm = rlm(abrasion.loss ~ hardness + ts.low, data = rubber, 
    psi = psi.bisquare)
```

-----

To visualize the fitted model, we need to get fitted values from the model on a grid of values of the two predictors.

```{r}
library(broom)
rubber.grid = expand.grid(hardness = c(54, 64, 74, 84),
                          tensile.strength = c(144, 162, 180, 198)) %>% data.frame
rubber.grid = rubber.grid %>% mutate(ts.low = tslow(tensile.strength))
rubber.predict = augment(rubber.rlm, newdata = rubber.grid)
rubber.predict
```

-----

Once we have the fitted values, we can make a coplot of the fitted model. We'll start with `hardness` as the given variable:
```{r}
ggplot(rubber.predict) +
    geom_line(aes(x = tensile.strength, y = .fitted)) +
    facet_grid(~ hardness) +
    scale_y_continuous("Fitted abrasion loss") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggplot(rubber.predict) +
    geom_line(aes(x = tensile.strength, y = .fitted, color = factor(hardness, ordered = TRUE))) +
    guides(color = guide_legend(title = "Hardness")) +
    scale_y_continuous("Fitted abrasion loss") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Note that the first plot is a coplot, the second doesn't have a name but reports the same information in a different way.

-----

Then a coplot with `tensile.strength` as the given variable:
```{r}
ggplot(rubber.predict) +
    geom_line(aes(x = hardness, y = .fitted)) +
    facet_grid(~ tensile.strength) +
    scale_y_continuous("Fitted abrasion loss") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggplot(rubber.predict) +
    geom_line(aes(x = hardness, y = .fitted, color = factor(tensile.strength, ordered = TRUE))) +
    guides(color = guide_legend(title = "Tensile strength")) +
    scale_y_continuous("Fitted abrasion loss") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


## Residuals

```{r}
rubber.resid = data.frame(rubber, .resid = residuals(rubber.rlm))
ggplot(rubber.resid, aes(x = tensile.strength, y = .resid)) + geom_point() + 
    stat_smooth(method = "loess", span = 1, method.args = list(degree = 1, family = "symmetric")) + 
    geom_abline(slope = 0, intercept = 0) +
    scale_y_continuous("Abrasion loss residuals")
```

-----

```{r}
ggplot(rubber.resid, aes(x = hardness, y = .resid)) + geom_point() + 
    stat_smooth(method = "loess", span = 1, method.args = list(degree = 1, family = "symmetric")) + 
    geom_abline(slope = 0, intercept = 0) +
    scale_y_continuous("Abrasion loss residuals")
```

## Coplots of the residuals

```{r}
resid_co_hardness = make_coplot_df(rubber.resid, faceting_variable = "hardness", number_bins = 4)
ggplot(resid_co_hardness, aes(x = tensile.strength, y = .resid)) +
    geom_point() +
    facet_grid(~ interval) +
    stat_smooth(method = "loess", method.args = list(degree = 1, family = "symmetric")) +
    scale_y_continuous("Abrasion loss residuals")
```


```{r}
resid_co_ts = make_coplot_df(rubber.resid, faceting_variable = "tensile.strength", number_bins = 4)
ggplot(resid_co_ts, aes(x = hardness, y = .resid)) +
    geom_point() +
    facet_grid(~ interval) +
    stat_smooth(method = "loess", method.args = list(degree = 1, family = "symmetric")) +
    scale_y_continuous("Abrasion loss residuals")
```


## Second-round model

From the residual plots, it looks like we might actually do better fitting an interaction between `tensile.strength` and hardness.

Exercises:

> - Refit a linear model that predicts `abrasion.loss` using an interaction between our non-linear transformation of `tensile.strength` and `hardness` (i.e., change `abrasion.loss ~ ts.low + hardness` to `abrasion.loss ~ ts.low * hardness`).

> - Plot the fitted values from the interaction model on the same grid of predictor variables we used in the no-interaction model. How does the form of the fits change when you add the interaction? Why is this?

> - Make residual plots and coplots for the interaction model in the same way we did for the no-interaction model. Do you like this model better?

## Final plots

```{r}
rubber.rlm = rlm(abrasion.loss ~ hardness * ts.low, data = rubber, psi = psi.bisquare)
coplot_hardness = coplot_hardness %>% mutate(interval_mean = get_interval_means(interval))
rubber.grid.final = expand.grid(hardness = unique(coplot_hardness$interval_mean),
                          tensile.strength = c(125, 180, 240)) %>% data.frame
rubber.grid.final = rubber.grid.final |> mutate(ts.low = tslow(tensile.strength))
rubber.predict.final = augment(rubber.rlm, newdata = rubber.grid.final)
rubber.predict.final = merge(rubber.predict.final,
                             unique(coplot_hardness[,c("interval", "interval_mean")]),
                             by.x = "hardness", by.y = "interval_mean")

ggplot(coplot_hardness, aes(x = tensile.strength, y = abrasion.loss)) +
    geom_point() +
    geom_line(aes(x = tensile.strength, y = .fitted), data = rubber.predict.final) +
    facet_wrap(~ interval, ncol = 4) +
    scale_y_continuous("Abrasion loss") +
    scale_x_continuous("Tensile strength")
```
