% Stat 470/670 Lecture 22: Logistic regression with many predictors
% Julia Fukuyama
% April 6, 2021

```{r setup, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(fig.cap="", fig.width = 5, fig.height = 3, dpi=175, fig.path="lecture-22-fig/", warning = FALSE)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(dplyr)
library(viridis)
library(broom)
set.seed(0)
```

# Arsenic in wells

The file `wells.dat` contains data on 3,020 households in Bangladesh whose wells had arsenic levels above the national drinking water standard of 50 micrograms per liter. The variables are:

- `switch`: the response: whether or not the household switched after
being encouraged to do so (after high arsenic was measured.)

- `arsenic`: the arsenic concentration in hundreds of micrograms per
liter.

- `dist`: the distance to the nearest safe well in meters.

- `assoc`: whether anyone in the household is in a community
association.

- `educ`: the years of education of the head of the household.

-----

We load at the data and look at the numerical summary and a pairs plot:

```{r, message = FALSE}
library(GGally)
wells = read.table("../../datasets/wells.dat")
summary(wells)
ggpairs(wells)
```

As we might expect, distance and education level are positively
associated with switching, while arsenic level is negatively
associated. Surprisingly, the correlation between belong to an
association and switching is negative, though the relationship is very
weak.

-----

Distance and arsenic levels should be the best predictors, so let's
look more closely at the distributions of those variables.

```{r}
ggplot(wells, aes(x = dist)) +
    geom_histogram(breaks = seq(0, 340, 10)) +
    xlab("Distance to nearest safe well (meters)")
```

We see the distribution of distance peaks around 40--60 meters, then
is strongly right-skewed. We should keep the possibility of a
transformation in mind.

-------

Then arsenic:

```{r}
ggplot(wells, aes(x = arsenic)) +
    geom_histogram(breaks = seq(0, 10, 0.25)) + xlab("Arsenic concentration (100 mg/liter)")
```

There are no observations with arsenic below 0.5, because these are
considered "safe" and the households don't get advised to switch
wells. Otherwise, the distribution is against strongly right-skewed,
suggesting a transformation. 

Note that the scales of `dist` and `arsenic` are quite
different. Some authors suggest standardizing the data for
interpretability, so that predictors are on approximately the same
scale. That's not a bad idea but we will hold off for now.

------

And finally education:

```{r}
ggplot(wells, aes(x = educ)) + geom_histogram(breaks = -0.5:17.5) +
    xlab("Years of education")
```

The distribution is weird. 0 (no education) and 5 (primary school only) are the magic numbers.

# Starting off simple

One way of approaching EDA is to throw everything into your model and
then get rid of terms you don't need. Another strategy is to start
with simple models and gradually build up to complex models. Let's
take this second approach for now. 

We'll start by predicting switching, using distance as the only
response. There's no reason why we can't try loess: 

```{r}
ggplot(wells, aes(x = dist, y = switch)) +
    geom_jitter(width = 0, height = 0.1) +
    geom_smooth(method = "loess") +
    xlab("Distance from nearest safe well") +
    ylab("Probability of switching")
```

We can interpret the fit as the probability of switching wells, given
the distance to the nearest safe well. There's a bump for very low
distances (which is probably just noise) and then a decline. We could
play around with the smoothing parameters to get something more
pleasant-looking. Instead, we'll fit a logistic regression to
guarantee a decreasing relationship.

-----

```{r}
gg = ggplot(wells, aes(x = dist, y = switch)) +
    geom_jitter(width = 0, height = 0.1) +
    geom_smooth(method = "loess")
gg + geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "orange") +
    xlab("Distance from nearest safe well") +
    ylab("Probability of switching")
```

The parametric method imposes a functional form on the data. This aids
in interpretability: the fit can now be described in a couple of
parameters, and the weird bump at the beginning is gone. The cost is
some lack of fit: perhaps the probability should decrease more quickly
as the distance gets beyond about 60 meters (remember that water is
heavy and it's tiring to carry buckets back and forth over long
distances.)

-----

Let's try the same kind of approach using arsenic concentration as the sole predictor.

```{r}
gg = ggplot(wells, aes(x = arsenic, y = switch)) +
    geom_jitter(width = 0, height = 0.1) +
    geom_smooth(method = "loess")
gg + geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "orange") +
    xlab("Arsenic concentration") +
    ylab("Probability of switching")
```

The fits diverge at arsenic levels above about 5. The seems to be
because of a couple of extremely high arsenic observations which led
to switching. Note however that not much of the data has arsenic above
5, so it remains to be seen how much this matters.

Also note the difference in fits for low levels of arsenic. This is
more worrisome than the difference at high levels because there is
actually a lot of data there. It seems likely that there is a lack of
fit for the low concentrations that logistic regression can't
handle. We'll come back to this later.

# Two predictors

We now want to know how the chance of switching depends on distance
and arsenic simulataneously. Before fitting a model, we plot the data
to get a feel for it. We'll use color to distinguish between
households that switch and households that don't. 

```{r}
ggplot(wells, aes(x = dist, y = arsenic, color = factor(switch))) +
    geom_point(alpha = 0.5) +
    xlab("Distance from nearest safe well") +
    ylab("Arsenic concentration") +
    labs(color = "Did they switch?") +
    scale_color_manual(values = c("#E69F00", "#56B4E9"), labels = c("No", "Yes"))
```

The `alpha = 0.5` makes the points slightly transparent, which can
help visually when you have a lot of data. Still, it's hard to work
out how switching depends on the other two variables from this graph;
that's why we're fitting a model. The main thing to take home from the
graph is the lack of data in the top right: we have no observations at
all with both distance above 200 and arsenic above 4, so we should not
try to generalize to this region.

-----

Density plots are actually more useful for comparing the two
distributions, and we can use `geom_density2d` to make such a plot:

```{r}
ggplot(wells, aes(x = dist, y = arsenic, color = factor(switch))) +
    geom_density2d() +
    xlab("Distance from nearest safe well") +
    ylab("Arsenic concentration") +
    labs(color = "Did they switch?") +
    scale_color_manual(values = c("#E69F00", "#56B4E9"), labels = c("No", "Yes"))
```

We see here that people who didn't switch tend to be at lower arsenic
levels and larger distances from the nearest safe well, which makes sense.

-----

Now we'll fit a logistic regression using both distance and arsenic as
predictors. We first fit an additive model, i.e. with no interaction.

```{r}
switch.logit = glm(switch ~ dist + arsenic, family = "binomial", data = wells)
summary(switch.logit)
```

As always, the coefficients are the most important things here. According to the model,

$$
\textrm{logit}[P(\textrm{switch}|{\textrm{dist, arsenic}})] = 0.003 - 0.00897 \times \textrm{dist} + 0.4608 \times \textrm{arsenic}.
$$

-----

As in the continuous case, we visualize the fit by drawing multiple
curves representing different values of one of the predictors. Let's
display the switching probability as a function of distance for a few
values of arsenic. Note that when we use the `augment()` function, we
specify `type.predict = "response"` to get the probabilities and not their
logits. 

```{r}
dist_df = expand.grid(dist = 0:339, arsenic = seq(0.5, 2.5, 0.5))
dist_preds = augment(switch.logit, type.predict = "response", newdata = dist_df)
ggplot(dist_preds, aes(x = dist, y = .fitted, group = arsenic, color = arsenic)) +
    geom_line() +
    xlab("Distance from nearest safe well") +
    ylab("Probability of switching") +
    labs(color = "Arsenic concentration") +
    scale_color_viridis()
```

The median arsenic level is 1.3. If arsenic is near the median, the
probability of switching declines from 60-something percent if you're
right by a safe well to less than 10% if the nearest safe well is
hundreds of meters away.

-----

Now find prediction curves for arsenic for a few different distances:

```{r}
arsenic_df = expand.grid(arsenic = seq(0.5, 10, 0.01), dist = seq(0, 100, 25))
arsenic_pred = augment(switch.logit, type.predict = "response", newdata = arsenic_df)
ggplot(arsenic_pred, aes(x = arsenic, y = .fitted, group = dist, color = dist)) +
    geom_line() +
    xlab("Arsenic concentration") +
    ylab("Probability of switching") +
    labs(color = "Distance from nearest safe well") + scale_color_viridis()
```


The median distance to a safe well is about 37 meters. At that
distance, if the arsenic level is only just over the safety threshold,
it's about 50--50 whether a household switched. On the other hand, at
the highest levels of arsenic, households will almost certainly switch
even if the nearest safe well is quite far.

# Adding an interaction

We now add an interaction term between distance and arsenic.

```{r}
switch_int = glm(switch ~ dist * arsenic, family = "binomial", data = wells)
summary(switch_int)
```

The numbers are a bit hard to interpret. For example, the sign of the
interaction term is negative, but it's hard to know exactly what this
means (especially since the signs for distance and arsenic go in
different directions.) The interpretation is easier if we just plot
curves.

-----

```{r}
dist_int = augment(switch_int, type.predict = "response", newdata = dist_df)
ggplot(dist_int, aes(x = dist, y = .fitted, group = arsenic, color = arsenic)) +
    geom_line() +
    xlab("Distance from nearest safe well") +
    ylab("Probability of switching") +
    labs(color = "Arsenic concentration") + scale_color_viridis()
```

The interaction brings the curves together as distance increases. If
the nearest safe well is close, it makes a big difference whether the
arsenic concentration is just over the limit or much bigger. If the
nearest safe well is far, it makes little difference: people are
unlikely to switch no matter the concentration. The curves meet up
beyond 300 meters, though we only have three observations where the
distance exceeds 300 meters.

-----

```{r}
arsenic_int_pred = augment(switch_int, type.predict = "response", newdata = arsenic_df)
ggplot(arsenic_int_pred, aes(x = arsenic, y = .fitted, group = dist, color = dist)) +
    geom_line() +
    xlab("Arsenic concentration") +
    ylab("Probability of switching") +
    labs(color = "Distance from nearest safe well") + scale_color_viridis()
```

The curves actually get further apart at first as arsenic increases
(up to a point.) That is, the curves for short distances rise quite
quickly as arsenic increases, while the curves for long distances rise
more slowly. Eventually, for exceptionally high levels of arsenic, the
curves come together (because they can't go any higher than 1.)

# Lots of predictors

Distance and arsenic are both good predictors and there's no reason to
think there shouldn't be an interaction, so keep the interaction
term. Now let's add `assoc` and `educ` to the model and see if the fit
makes sense.

```{r}
summary(glm(switch ~ dist * arsenic + assoc + educ, family = "binomial", data = wells))
```

Years of education has a positive coefficient, which makes sense. We
can keep that in the model.

Association membership has a *negative* coefficient, which doesn't
make causal sense: it would be strange for association membership to
make you *less* likely to switch wells, unless you were a member of
the Association for Arsenic Being Good for You. If our sole goal was
prediction, we could do a careful cross-validation to see if
association membership did help predict more accurately. Since we're
doing EDA, we can drop the association term.

```{r}
summary(glm(switch ~ dist * arsenic + educ, family = "binomial", data = wells))
```

----

We now add the remaining two-way interactions. In general, if you only
have a few terms, you might as well include all the two-way
interactions unless you have a good reason not to.

```{r}
switch_model = glm(switch ~ dist + arsenic + educ + dist:arsenic + dist:educ + arsenic:educ,
    family = "binomial", data = wells)
```

-----

At this point our model is too complicated for it to be worth trying
to interpret the numbers. Let's extract the fitted values and
residuals, and plot the latter against the former. Note that the
default of `augment()` is to extract the deviance residuals instead of
the response residuals, if you know about and prefer deviance
residuals you can specify them instead.

```{r}
switch_model_df = augment(switch_model, type.residuals = "pearson")
ggplot(switch_model_df, aes(x = .fitted, y = .resid)) +
    geom_point() +
    geom_smooth(method = "loess", method.args = list(degree = 1)) +
    xlab("Fitted values") +
    ylab("Residuals")
```

The points on the plot fall exactly on two lines, so look
at the smooth instead. We see there's a little bit of waviness. This
is fairly typical for logistic regression: there's no reason why the
relationship between probability and the predictors should have that
exact functional form. But the improvement in fit you get from fitting
a nonparametric model is fairly small.

-----

Now plot the residuals against distance. What we're looking for here
is whether the function form is correct, or whether we need more
flexibility in how we use distance to predict switching.

```{r}
ggplot(switch_model_df, aes(x = dist, y = .resid)) +
    geom_point() +
    geom_smooth(method = "loess", method.args = list(degree = 1)) +
    xlab("Distance from nearest safe well") +
    ylab("Residuals")
```

There's a wiggle, but this is basically fine -- the zero line is
enclosed in the confidence band.

-----

Now let's do the same plot with arsenic on
the $x$-axis:

```{r}
ggplot(switch_model_df, aes(x = arsenic, y = .resid)) +
    geom_point() +
    geom_smooth(method = "loess", method.args = list(degree = 1)) +
    xlab("Arsenic concentration") +
    ylab("Residuals")
```

This is much worse. The curve is too low, then rises too high, then
declines again. In fact, the extreme left hand side is the biggest
problem, since the lack is fit is real and the data is dense
there. The negative average residual there means that for arsenic
levels just over the threshold, the probability of switching is
*overestimated* (negative average residuals occur when there are more
zeroes in the response than the model expects.)

-----

Education isn't that interesting since the effect is much smaller, but
let's do it for completeness.

```{r}
ggplot(switch_model_df, aes(x = educ, y = .resid)) +
    geom_point() +
    geom_smooth(method = "loess", method.args = list(degree = 1)) +
    xlab("Years of education") +
    ylab("Residuals")
```

Again, the fit is imperfect.

-----

Remember when we said we might need to transform? The last couple of
graphs suggest that while a transformation of distance is unnecessary,
a transformation of arsenic is.  A transformation of education would
probably help too, but since the variable is weird and not that
important we won't bother. We'll take the log arsenic and refit the
model.

```{r}
wells$log.arsenic = log(wells$arsenic)
log_arsenic_model = glm(switch ~ dist + log.arsenic + educ + dist:log.arsenic + dist:educ + log.arsenic:educ, family = "binomial", data = wells)
log_arsenic_model_df = augment(log_arsenic_model, type.residuals = "pearson", type.predict = "response")
```

-----

Now plot the residuals of this new model against log arsenic:

```{r}
ggplot(log_arsenic_model_df, aes(x = log.arsenic, y = .resid)) +
    geom_point() +
    geom_smooth(method = "loess", method.args = list(degree = 1))  +
    xlab("Log arsenic concentration") +
    ylab("Residuals")
```

It's a lot better. I'm fairly happy with the model at this
point. 

-----

Let's take a look at the fit. Fix education at its median, and
draw arsenic curves for several distances:

```{r}
switch_grid = expand.grid(arsenic = seq(0.5, 10, 0.01), dist = seq(0, 100, 25), educ = 5)
switch_grid$log.arsenic = log(switch_grid$arsenic)
log_arsenic_pred_grid = augment(log_arsenic_model, newdata = switch_grid,
    type.predict = "response", type.residuals = "pearson")
ggplot(log_arsenic_pred_grid, aes(x = arsenic, y = .fitted, group = dist, color = dist)) +
    geom_line() +
    xlab("Arsenic concentration") +
    ylab("Probability of switching") + scale_color_viridis()
```

-----

There are 3! ways you can assign the three variables to $x$-axis,
conditioning variable, and fixed. You probably get the idea at this
point, though, so let's just do one more:

```{r}
switch_grid_2 = expand.grid(arsenic = seq(0.5, 10, 0.01), dist = median(wells$dist), educ = c(0, 5, 10))
switch_grid_2$log.arsenic = log(switch_grid_2$arsenic)
log_arsenic_pred_grid_2 = augment(log_arsenic_model, newdata = switch_grid_2,
    type.predict = "response", type.residuals = "pearson")
ggplot(log_arsenic_pred_grid_2,
       aes(x = arsenic, y = .fitted, group = educ, color = factor(educ))) +
    geom_line() +
    xlab("Arsenic concentration") +
    ylab("Probability of switching") +
    labs(color = "Education") +
    scale_color_manual(values = c("#999999", "#E69F00", "#56B4E9"),
                       labels = c("0 years", "5 years", "10 years"))
```


# Improving the model

If you really care about prediction but still want to be able to
interpret the model, try a nonparametric model such as a generalized
additive model (GAM) or loess. To overgeneralize, GAM is often
better/easier when you have lots of predictors, while loess is often
better/easier when you have complex interactions. You can learn more
about these in S425/625.

If you really, really care about predictions then you can use machine
learning techniques. The improvement in prediction is usually quite
small, however, and you'll lose a lot or all the interpretability.

